---
title: "KSPZV1 Correlate Cellular and CSP IgG Features to Time to Parasitemia through 6 months"
author: "Tuan M. Tran"
date: "09/27/2021"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(googledrive)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(fgsea)
library(data.table)
library(Biobase)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective

Plot correlations between adaptive features (both baseline and post-vax) and time to first Pf infection through 6 months of surveillance post-vaccination.

### Load data

```{r load data, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("1pSt3bzf_F8KO4mQRQojLuYHexGewLwPI"), path = temp, overwrite = TRUE)
summarized_correlations <- readRDS(file = dl$local_path)
```

### CSP IgG reactivity at baseline and post-vax by treatment and outcome (Fig 3D of pre-print)

```{r, volcano plot, echo=TRUE, warning=FALSE, message=FALSE}
cols <- c("baseline" = "blue", "2 weeks post-vax" = "red", "log2 fold-change" = "orange")
cols2 <- c("baseline" = "blue", "2 weeks post-vax" = "red", "log2 fold-change" = "orange", "not significant" = "gray")
myPlot <- ggplot(summarized_correlations, aes(x = pearson_r, y = neglogpval, fill = timepoint2, color = timepoint, label = label)) +
  geom_point(shape = 21, alpha = 0.75, size = 2) +
  ggrepel::geom_text_repel(aes(label=label),hjust=0, vjust=0, color = "black", force_pull = 1, size = 3.25, segment.size = 0.2) +
  scale_color_manual(values = cols, aesthetics = "colour") +
  scale_fill_manual(values = cols2, aesthetics = "fill") +
  theme_bw(base_family = "Arial", base_size = 14) +
  geom_hline(yintercept = 1.3, linetype = "dotted", color = "red")+
  ylab("-log10(p value)") +
  xlab("Pearson's r") +
  xlim(c(-0.4,0.4)) +
  theme(legend.position = "top", legend.box="vertical", legend.margin=margin())
```

Volcano plot of CSP-specific IgG and flow cytometry features at each timepoint or calculated as fold-change
post-vaccination over baseline. Red dashed line indicates P = 0.05.

```{r, print volcano plot, echo=FALSE, fig.align='center', fig.width=6.5, fig.height=6, message=FALSE, warning=FALSE}
print(myPlot)
```

### Show correlation between CSP-specific memory B cells and time-to-first parasitemia up to 6 months as scatter plot (Fig 3B of pre-print)

Correlation between CSP-specific memory B cells and time-to-first parasitemia up to 6 months (n=183) by presence (not protected, NP) or absence (protected, P) of parasitemia through 3 months of follow up

```{r plot CSP-specific of memory B cells vs TTE 6 months, message=FALSE, warning=TRUE}
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("1v2l7G7AzXhgcqObzAabYy6_QJsdB5hFn"), path = temp, overwrite = TRUE)
longdat.tte <- readRDS(file = dl$local_path)

longdat.tte.CSPmemB <- longdat.tte %>%
  dplyr::filter(feature == "CSP-specific of memory B cells") %>%
  dplyr::filter(!is.na(Post_Vaccination)) %>%
  dplyr::filter(!is.na(tte.mal.atp.6)) 
CSPmemB_TTE_plot <- ggplot(longdat.tte.CSPmemB, aes(x = tte.mal.atp.6, y = Post_Vaccination)) +
  geom_point() +
  theme_bw()

myCorMeth <- "pearson"
CSPmemB_TTE_plot <- ggscatter(longdat.tte.CSPmemB, x = "tte.mal.atp.6", y = "Post_Vaccination",
            conf.int = TRUE , add = "reg.line", size = 2,
            cor.coef.size = 1,
            add.params = list(color = "black", fill = "lightgray", alpha = 0.4, size = 1),
            fill = "Outcome", color = "Outcome") +
  stat_cor(method = myCorMeth) +
  scale_fill_manual(values = c("#A6CEE3", "#1F78B4")) +
  scale_color_manual(values = c("#A6CEE3", "#1F78B4")) +
  xlab("days to first parasitemia (6 month follow-up)") +
  ylab("% CSP-specific of memory B cells") +
  theme_bw(base_family = "Arial", base_size = 14) +
  theme(legend.position="top") 
```

```{r,  print scatter plot, echo=FALSE, fig.align='center', fig.width=5, fig.height=4.25, message=FALSE, warning=FALSE}
print(CSPmemB_TTE_plot)
```

### Correlate Baselinelines with CSP-specific memory B cells

```{r prepare cpm data, warning=FALSE, message=FALSE}
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("1togaBlNIxiDj16FyXTC-r7Qw0A9cG2az"), path = temp, overwrite = TRUE)
x  <- readRDS(file = dl$local_path)
```

```{r correlate baseline genes with CSP-specific, message=FALSE, warning=FALSE}
csp.memoryB.dat <- longdat.tte %>%
  dplyr::select(PATID, Dosegroup, Outcome, feature, Post_Vaccination) %>%
  dplyr::filter(feature == "CSP-specific of memory B cells") %>%
  pivot_wider(names_from = "feature", values_from = "Post_Vaccination")
  
cpm.dat <- exprs(x) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "SAMPLEID") %>%
  dplyr::filter(grepl("_0", SAMPLEID)) %>%
  mutate(PATID = gsub("_0", "", SAMPLEID)) %>%
  dplyr::select(PATID, everything()) %>%
  dplyr::select(-c(SAMPLEID)) %>%
  pivot_longer(cols = 2:ncol(.), names_to = "EnsemblID", values_to = "CPM") %>%
  left_join(., csp.memoryB.dat,
            by = "PATID") %>%
  left_join(., fData(x) %>%
               dplyr::select(EnsemblID, GeneSymbol),
             by = "EnsemblID")

summarized_cpm_CSPmemB <- cpm.dat %>%
  dplyr::filter(!is.na(`CSP-specific of memory B cells`)) %>%
  group_by(GeneSymbol, EnsemblID) %>%
  summarise(baselineCPM_pearson_r_CSPMemB = stats::cor.test(CPM, `CSP-specific of memory B cells`, method = myCorMeth)$estimate,
            baselineCPM_pearson_pval_CSPMemB = stats::cor.test(CPM, `CSP-specific of memory B cells`, method = myCorMeth)$p.val) %>%
  mutate(RankMetric = sign(baselineCPM_pearson_r_CSPMemB)*-(log10(baselineCPM_pearson_pval_CSPMemB))) %>%
  arrange(desc(RankMetric))
```

### Apply GSEA

```{r apply fgsea, warning=FALSE, message=FALSE}
res <- summarized_cpm_CSPmemB %>%
  dplyr::select(GeneSymbol, RankMetric) %>%
  na.omit() %>% 
  distinct() %>% 
  group_by(GeneSymbol) %>% 
  summarize(RankMetric = mean(RankMetric)) %>%
  arrange(desc(RankMetric)) 
ranks <- deframe(res)
head(ranks, 20)
#ignore warning about ties since there is only one duplicated RankMetric ranks[DescTools::AllDuplicated(ranks)]
#AC069547.1 RP11-521C22.3 
#0.0001374268  0.0001374268 
ranks <- deframe(res)

devtools::source_url("https://github.com/TranLab/ModuleLists/blob/main/NamedGeneRankList2GseaTable.R?raw=TRUE")
bound_df_GSEA <- NamedGeneRankList2GseaTable(rankedgenes = ranks, geneset = "bloodmodules", output_directory = tempdir(), filename_prefix = "GSEA_baseline_CPM_postvax_CSPmemB_AllGroups", minSize = 20, fixed_seed = TRUE)
```

### Plot GSEA results as bar plot (Fig 3C of pre-print)

Gene set enrichment analysis (GSEA) using genes ranked by magnitude and significance of correlation between baseline expression and % CSP-specific of memory B cells at 2-weeks post-vaccination.

```{r arrange data for plots, echo = FALSE}
myfunction <- function(leadingEdge) { length(unlist(strsplit(leadingEdge, ", "))) }
myFDRcutoff <- 0.05

myModules <- c("lowBTMs", "highBTMs", "MonacoModules", "BloodGen3Module", "MSigDB_Hallmark_v7.4")
myGSEAClusterPlotDat <- bound_df_GSEA %>%
  filter(padj < myFDRcutoff) %>%
  filter(module_type %in% myModules) %>%
  dplyr::mutate(edges = purrr::pmap_dbl(list(leadingEdge), myfunction)) %>%
  dplyr::mutate(bubblelabel = paste(edges,size, sep = "/")) %>%
  group_by(module_type) %>%
    arrange(module_type, desc(NES), padj) %>%
    dplyr::select(module_type, pathway, edges, size, bubblelabel, NES, padj) %>%
  ungroup() %>%
  filter(!grepl("TBA", pathway)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = gsub("Vd2", "Vδ2", pathway)) %>%
  mutate(pathway = gsub("Vg", "Vγ", pathway)) %>%
  mutate(pathway = gsub("HALLMARK_", "", pathway)) %>%
  mutate(pathway = gsub("_", " ", pathway)) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  arrange(neglogpadj) %>%
  mutate(pathway = fct_reorder(pathway, NES, .desc=TRUE)) %>%
  mutate(pathway = fct_reorder(pathway, neglogpadj)) %>%
  mutate(TextLabelColor = ifelse(module_type == "lowBTMs", scales::muted("red"),
                                 ifelse(module_type == "highBTMs", scales::muted("blue"),
                                        ifelse(module_type == "MonacoModules", "black","orange")))) %>%
  arrange(desc(neglogpadj))
myColors <- rev(myGSEAClusterPlotDat$TextLabelColor)
myGSEAClusterPlot <- myGSEAClusterPlotDat %>%
  ggplot(., aes(x = NES, y = pathway, fill = neglogpadj)) +
  geom_bar(stat = 'identity') + 
  viridis::scale_fill_viridis(option= "A", begin = 0.25, end = 0.75, alpha = 0.8, direction = -1, name = "neglogpadj") +
  theme_classic(base_family = "Arial", base_size = 14) +
  theme(legend.position = "bottom", axis.text.y = element_text(colour = myColors))
```

```{r print GSEA plot, fig.align='center', fig.height=6, fig.width=6, echo=FALSE}
addSmallLegend <- function(myPlot, pointSize = 6, textSize = 6, spaceLegend = 1) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize))) +
        theme(legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"))
}
print(addSmallLegend(myGSEAClusterPlot))
```