---
title: "KSPZV1 WGCNA baseline"
author: "Tuan M. Tran"
date: "09/12/2021"
output: html_document
---

### Objective

Perform weighted gene correlation network analysis as originally described by Horvath et al. on pre-immunization baseline transcriptomic data from the KSPZV1 clinical trial.

References:

https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/
https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/JMiller/Tutorial%20document.pdf

### Load packages

```{r load packages, message=FALSE, warning=FALSE}
library(edgeR)
library(readxl)
library(EDASeq)
library(Biobase)
library(WGCNA)
library(ape)
library(Cairo)
library(CorLevelPlot)
library(tidyverse)
library(igraph)
library(remotes)
library(fgsea)
library(data.table)
library(ggplot2)
library(viridis)
library(ggpubr)
library(googledrive)
allowWGCNAThreads()
```

### Options and define variables

```{r options and define variables, message=FALSE, warning=FALSE}
myCor = "bicor"
power <- 12.5 #determined by evaluating previous plot using pickSoftThreshold
myMergeCutHeight <- 0.05
myDeepSplit <- 2
minModSize <- 20
enforceMMS <- FALSE
cor.pval <- 0.05
```

### Load ExpressionSet

```{r load cpm eset, message=FALSE, warning=FALSE}
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("1togaBlNIxiDj16FyXTC-r7Qw0A9cG2az"), path = temp, overwrite = TRUE)
x <- readRDS(file = dl$local_path)
dim(x)
```

### Make weighted gene correlation matrix based on full data set

```{r weighted gene correlation matrix based on full data set, message=FALSE, warning=FALSE}
WGCNA_matrix <- t(exprs(x)) #make correlations only with full eset
blockSize(ncol(WGCNA_matrix), rectangularBlocks = TRUE, maxMemoryAllocation = 4^31)
par(mfrow=c(1,1))
plotClusterTreeSamples(datExpr=WGCNA_matrix)
#used bicor for Timepoint 0
powers <- seq(4,20,by=0.5)
sft <- pickSoftThreshold(WGCNA_matrix, powerVector = powers, corFnc = myCor, verbose = 5, networkType ="signed", blockSize = ncol(WGCNA_matrix))
sft$powerEstimate
par(mfrow=c(1,1))
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab='Soft Threshold (power)',ylab='Scale Free Topology Model Fit,signed R²',
     type='n', main = paste('Scale independence'));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=1,col='red'); abline(h=0.90,col='red')
```

### Run an automated network analysis

```{r calculation of adjacency matrix, warning=FALSE, message=FALSE}
#use blockwiseModules given many genes
net <- blockwiseModules(WGCNA_matrix,
                        power=power,
                        deepSplit= myDeepSplit,
                        minModuleSize=minModSize,
                        TOMType="none", 
                        mergeCutHeight=myMergeCutHeight, 
                        TOMDenom="mean",
                        detectCutHeight=0.995,
                        corType=myCor,
                        networkType="signed",
                        pamStage=TRUE,
                        pamRespectsDendro=TRUE,
                        reassignThresh=0.05,
                        verbose=5,
                        saveTOMs=FALSE,
                        maxBlockSize=ncol(WGCNA_matrix), 
                        nThreads = 0)

nModules <- length(table(net$colors))-1
modules <- cbind(colnames(as.matrix(table(net$colors))),table(net$colors))
orderedModules <- cbind(Mnum=paste("M",seq(1:nModules),sep=""),Color=labels2colors(c(1:nModules)))
modules <- modules[match(as.character(orderedModules[,2]),rownames(modules)),]
tmpMEs <- MEs <- net$MEs
colnames(tmpMEs) <- paste("ME",colnames(MEs),sep="")
kMEdat <- signedKME(WGCNA_matrix, tmpMEs, corFnc=myCor) #calculate (signed) eigengene-based connectivity, also known as module membership
WGCNA_dat <- cbind(fData(x)$GeneSymbol, colnames(WGCNA_matrix),net$colors,kMEdat) %>%
  as.data.frame() %>%
  dplyr::rename(GeneSymbol = "fData(x)$GeneSymbol") %>%
  dplyr::rename(ENSEMBLID = "colnames(WGCNA_matrix)") %>%
  dplyr::rename(ModuleColors = "net$colors")
```

### Correlate modules with traits

```{r correlate modules with traits, message=FALSE, warning=FALSE}
# Define numbers of genes and samples
nGenes = ncol(WGCNA_matrix)
nSamples = nrow(WGCNA_matrix)

datvar <- pData(x) %>%
  dplyr::select(PATID, SEQBATCH, site, SEX, age.vax1, mal.vax.1, treat, mal.atp.3, tte.mal.atp.6, mal.dvax, mal.dvax.tot, pfcsp_pre, pfcsp_post, log2FC_CSPAb) %>%
  mutate('Gender (female)' = factor(ifelse(SEX == "F", 1, 0))) %>%
  mutate('Site (Siaya)' = factor(ifelse(site == "Siaya", 1, 0))) %>%
  dplyr::rename('Pf infection at first vax' = "mal.vax.1") %>%
  mutate('pre-vax anti-CSP IgG' = log2(pfcsp_pre+1)) %>%
  dplyr::rename('log2FC anti-CSP IgG' = "log2FC_CSPAb") %>%
  mutate('1.8 x 10^6 PfSPZ' = factor(ifelse(treat == "1.8 x 10^6 PfSPZ", 1, 0))) %>%
  mutate('parasitemic events during vax period' = mal.dvax.tot) %>%
  mutate('uninfected, 3 months' = factor(ifelse(mal.atp.3 == 0, 1, 0))) %>%
  dplyr::rename('days to first parasitemia' = "tte.mal.atp.6") %>%
  mutate(Age = age.vax1) %>%
  dplyr::select(PATID, Age, 'pre-vax anti-CSP IgG', 'Pf infection at first vax', 'uninfected, 3 months', 'days to first parasitemia', 'log2FC anti-CSP IgG') %>% 
  as_tibble() %>%
  column_to_rownames(var = "PATID") %>%
  mutate_if(is.factor, as.character) %>%
  mutate_if(is.character, as.numeric)

modTraitCor <- cor(orderMEs(net$MEs), datvar, use = "p")
modTraitP <- corPvalueStudent(modTraitCor, nSamples)
#Since we have a moderately large number of modules and traits, a suitable graphical representation will help in reading the table. We color code each association by the correlation value: Will display correlations and their p-values

#Select out only modules that have P<0.05 in Protection

modTraitP.temp <- modTraitP %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module") %>%
  filter(.$'uninfected, 3 months' < cor.pval | .$'days to first parasitemia' < cor.pval)
modTraitCor.select <- modTraitCor[modTraitP.temp$Module,]
modTraitP.select <- modTraitP[modTraitP.temp$Module,]
textMatrix <- paste(signif(modTraitCor.select, 2), "\n(P=",
                   signif(modTraitP.select, 1), ")", sep = "")
dim(textMatrix) <- dim(modTraitCor.select)
```

### Display the correlation values within a heatmap plot (Figure 2A of pre-print)

```{r plot module trait correlations heatmap, warning=FALSE, message=FALSE, fig.align='center', fig.width=8, fig.height=8}
par(mar = c(11, 9, 1, 1))
labeledHeatmap(Matrix = modTraitCor.select, xLabels = names(datvar),
               yLabels = rownames(modTraitCor.select), ySymbols = rownames(modTraitCor.select),
               colorLabels =FALSE,colors=blueWhiteRed(100),textMatrix=textMatrix,
               setStdMargins = FALSE, zlim = c(-1,1),
               main = paste("Module-trait relationships"),xLabelsAngle = 45) 
```

### Identify hub genes


```{r id hub genes, message=FALSE, warning=FALSE}
myColors <- gsub("ME", "", rownames(modTraitCor.select))
topmodules <- chooseTopHubInEachModule(WGCNA_matrix, net$colors, omitColors = "grey", power = 12.5, type ="signed")
mytopmodules <- topmodules[myColors] %>%
  as.data.frame() %>%
  dplyr::rename(EnsemblID = ".") %>%
  rownames_to_column("module_label") %>%
  as_tibble() %>%
  left_join(., fData(x) %>%
              dplyr::select(EnsemblID, GeneSymbol), by = "EnsemblID")
devtools::source_url("https://github.com/jtlovell/limmaDE2/blob/master/R/wgcna2igraph.R?raw=TRUE")
graph <- wgcna2igraph(net = net, WGCNA_matrix, modules2plot = myColors,
                      colors2plot =  myColors,
                      kME.threshold = 0.5, adjacency.threshold = 0.1,
                      adj.power = power, verbose = T,
                      node.size = 1.5, frame.color = NA, node.color = scales::muted("red"),
                      edge.alpha = .7, edge.width = 0.5)
hubscores <- hub_score(graph, scale = TRUE, weights = NULL,
  options = arpack_defaults)
```

### Top modules and hub genes (rownames for Figure 2A of pre-print)

```{r display my topmodules, echo=FALSE}
knitr::kable(mytopmodules)
```


### Plot network graph of significant modules (Figure 2B of pre-print)

Network graphs of significant modules containing nodes (red dots) and edges (lines) meeting minimum thresholds. Correlations between nodes in different modules are shown as black edges.

```{r plot networkd graph, message=FALSE, warning=FALSE, fig.align='center', fig.width=8, fig.height=8, echo=FALSE}
plot(graph)
```

### Identify modules with < 0.05

```{r id modules that are significant than pull out gene list, message=FALSE, warning=FALSE}
WGCNA_dat_select <- c()
for(i in modTraitP.temp$Module){
  module.color <- sub("ME","", i)
  module.colname <- paste0("kME", i)
  WGCNA_dat_select[[i]] <- WGCNA_dat %>% 
    filter(ModuleColors == module.color) %>%
    dplyr::select(GeneSymbol, all_of(module.colname))
}
lapply(WGCNA_dat_select, nrow)
```

### FGSEA on WGCNA Results

rank by module membership

```{r FGSEA on WGCNA Results, warning=FALSE, message=FALSE}
minSize <- 5
# Make rank list; ranked by ModuleMembership correlation
  ranks <- GSEA_baseline_bound_df <- c()
devtools::source_url("https://github.com/TranLab/ModuleLists/blob/main/NamedGeneRankList2GseaTable.R?raw=TRUE")
for(k in names(WGCNA_dat_select)){
  WGCNA_dat_select[[k]] <- WGCNA_dat_select[[k]][order(WGCNA_dat_select[[k]][,2], decreasing = TRUE),]
  ranks[[k]] <- WGCNA_dat_select[[k]][,2]
  names(ranks[[k]]) <- WGCNA_dat_select[[k]]$GeneSymbol
  GSEA_baseline_bound_df[[k]] <- NamedGeneRankList2GseaTable(rankedgenes = ranks[[k]], geneset = "all", output_directory = tempdir(),
                                                        filename_prefix = paste0("FGSEA_Mod_Corr_Protect_3_mos_", k,
                                                                                 "_minSize", minSize), scoreType = "pos", minSize = minSize, fixed_seed = TRUE)
  }
```

### Plot WGCNA GSEA Results

baseline analysis

```{r plot wgcna gsea, warning=FALSE, message=FALSE}
addSmallLegend <- function(myPlot, pointSize = 1.5, textSize = 3, spaceLegend = 0.3) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize))) +
        theme(legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"))
}
myModuleTypes <- c("MSigDB_Hallmark_v7.4", "MSigDB_C2_kegg_v7.4", "highBTMs", "lowBTMs", "BloodGen3Module")
myGSEAClusterPlotDat <- GSEA_baseline_bound_df$MElavenderblush2 %>%
  mutate(module_hub = "EFHD2") %>%
  bind_rows(., GSEA_baseline_bound_df$MEthistle3 %>% mutate(module_hub = "SEC62") ) %>%
    bind_rows(., GSEA_baseline_bound_df$MEmediumpurple1  %>%
  mutate(module_hub = "CSDE1")) %>%
  bind_rows(., GSEA_baseline_bound_df$MEskyblue1  %>%
  mutate(module_hub = "RIOK3")) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  mutate(pathway = gsub("VS", "v", pathway)) %>%
  mutate(pathway = gsub("Vd", "Vδ", pathway)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = sub(".*?\\_", "", pathway)) %>%
  mutate(pathway = fct_reorder(pathway, NES))  %>%
  arrange(desc(NES))%>%
  mutate(TextLabelColor = ifelse(module_type == "BloodGen3Module", scales::muted("red"),
                                 ifelse(module_type == "MSigDB_C2_kegg_v7.4", scales::muted("blue"),
                                        ifelse(module_type == "MSigDB_Hallmark_v7.4", "black","gray")))) %>%
  filter(padj < 0.10) %>%
  filter(module_type %in% myModuleTypes)
```

### Arrange plots using ggarrange

```{r make ggarranged plot, warning=FALSE, message=FALSE}

scale_begin <- min(myGSEAClusterPlotDat$neglogpadj, na.rm = TRUE)
scale_end <- max(myGSEAClusterPlotDat$neglogpadj, na.rm = TRUE)

myGSEAClusterPlot_EFHD2 <- myGSEAClusterPlotDat %>%
  filter(module_hub == "EFHD2") %>%
  ggplot(., aes(x = NES, y = pathway, fill = neglogpadj)) +
  geom_bar(stat = 'identity') +
  scale_fill_gradientn(colors = viridis::viridis_pal(option = "A", alpha = 0.8, direction = -1, begin = 0.25, end = 1)(9), limits=c(scale_begin, scale_end), name = "neglogpadj", na.value = "#FDE725FF") + 
  ylab("EFHD2") +
  xlim(c(0, max(myGSEAClusterPlotDat$NES))) +
  theme_classic(base_family = "sans", base_size = 6) +
  theme(legend.position = "bottom", axis.text.y = element_text(colour = rev(myGSEAClusterPlotDat$TextLabelColor)),
        plot.margin = unit(c(0,0,0,4), "cm"))

myGSEAClusterPlot_RIOK3 <- myGSEAClusterPlotDat %>%
  filter(module_hub == "RIOK3") %>%
  ggplot(., aes(x = NES, y = pathway, fill = neglogpadj)) +
  geom_bar(stat = 'identity') + 
  scale_fill_gradientn(colors = viridis::viridis_pal(option = "A", alpha = 0.8, direction = -1, begin = 0.25, end = 1)(9), limits=c(scale_begin, scale_end), name = "neglogpadj", na.value = "#FDE725FF") + 
  ylab("RIOK3") +
  xlim(c(0, max(myGSEAClusterPlotDat$NES))) +
  theme_classic(base_family = "sans", base_size = 6) +
  theme(legend.position = "none", axis.text.y = element_text(colour = rev(myGSEAClusterPlotDat$TextLabelColor)),
        plot.margin = unit(c(0,0,0,6), "cm"))

myGSEAClusterPlot_CSDE1 <- myGSEAClusterPlotDat %>%
  filter(module_hub == "CSDE1") %>%
  ggplot(., aes(x = NES, y = pathway, fill = neglogpadj)) +
  geom_bar(stat = 'identity') + 
  scale_fill_gradientn(colors = viridis::viridis_pal(option = "A", alpha = 0.8, direction = -1, begin = 0.25, end = 1)(9), limits=c(scale_begin, scale_end), name = "neglogpadj", na.value = "#FDE725FF") + 
  ylab("CSDE1") +
  xlim(c(0, max(myGSEAClusterPlotDat$NES))) +
  theme_classic(base_family = "sans", base_size = 6) +
  theme(legend.position = "none", axis.text.y = element_text(colour = rev(myGSEAClusterPlotDat$TextLabelColor)),
        plot.margin = unit(c(0,0,0,6), "cm"))

myGSEAClusterPlot_SEC62 <- myGSEAClusterPlotDat %>%
  filter(module_hub == "SEC62") %>%
  ggplot(., aes(x = NES, y = pathway, fill = neglogpadj)) +
  geom_bar(stat = 'identity') + 
  scale_fill_gradientn(colors = viridis::viridis_pal(option = "A", alpha = 0.8, direction = -1, begin = 0.25, end = 1)(9), limits=c(scale_begin, scale_end), name = "neglogpadj", na.value = "#FDE725FF") + 
  ylab("SEC62") +
  xlim(c(0, max(myGSEAClusterPlotDat$NES))) +
  theme_classic(base_family = "sans", base_size = 6) +
  theme(legend.position = "none", axis.text.y = element_text(colour = rev(myGSEAClusterPlotDat$TextLabelColor)))

myArrangedPlot <- ggarrange(ggarrange(myGSEAClusterPlot_EFHD2, myGSEAClusterPlot_EFHD2, nrow = 1, common.legend = TRUE),
                            ggarrange(ggarrange(myGSEAClusterPlot_SEC62, myGSEAClusterPlot_CSDE1, nrow = 2), myGSEAClusterPlot_RIOK3, ncol = 2,
                                             heights = c(1.5, 1)), nrow = 2, heights=c(5,2), widths=c(0.75, 1), common.legend = TRUE)

```

### GSEA bar plots (Figure 2C of pre-print)

GSEA of genes within modules that significant correlated with outcome using blood transcription modules, BloodGen3Modules, and KEGG pathways as gene sets (BH-adjusted p<0.10). The SEC62-hubbed module did not show significant enrichment for any gene set.

```{r print my arrange plot, fig.align='center', fig.width=10, fig.height=6, warning=FALSE, message=FALSE, echo=FALSE}
print(myArrangedPlot)
```
