---
title: "PfSPZ PfSPZ-Induced Features that Predict Protection by TTE"
author: "Tuan M. Tran"
date: "10/13/2021"
output: html_document
---

```{r setup, include=FALSE}
library(ComplexHeatmap)
library(tidyverse)
require(miscTools)
library(circlize)
library(gtools)
library(survival)
library(survminer)
```

### Objective

Use a rational approach to determine which features can distinguish between vaccine-induced protecte and not protected.

Criteria:

  1. Gene is differentially expressed in delta P vs delta S analysis in high-dose PfSPZ (p<0.005)
  2. Gene induced post-vax in high-dose PfSPZ protected (p<0.05)
  3. Gene downregulated post-vax in high-dose PfSPZ NOT protected (p<0.15)

### Read-in data and filter

```{r read in data, echo=TRUE, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- googledrive::drive_download(
  googledrive::as_id("1R8bLYmyp95h9NVtP6hmNqSwzi51ujKgf"), path = temp, overwrite = TRUE)
deltaDEGs <- readxl::read_excel(dl$local_path)
#1 
deltadelta <- deltaDEGs %>%
  filter(comparison == "ΔP vs ΔNP" & PValue < 0.005)
#2
deltaprotected <- deltaDEGs %>%
  filter(comparison == "ΔP" & logFC > 0 & PValue < 0.05)
#3
deltaprotectedNotprotected <- deltaDEGs %>%
  filter(comparison == "ΔNP" & logFC < 0 & PValue < 0.1)

allDat <- rbind(deltadelta, deltaprotected, deltaprotectedNotprotected) %>%
  select(hgnc_symbol, description, logFC, PValue, FDR, comparison) %>%
  pivot_wider(names_from = comparison, values_from = c(logFC, PValue, FDR)) %>%
  filter(complete.cases(.))

myGenes <- allDat$hgnc_symbol
```

### Table of filtered genes

Figure 4H in preprint

```{r table of filtered genes, echo = FALSE, message=FALSE, warning=FALSE}
knitr::kable(allDat[,c(1,2,3,6,4,7,5,8)])
```

```{r check nanostring list, echo = FALSE, eval=FALSE}
#get existing Nanostring list
nanostring <- readxl::read_excel("/Volumes/GoogleDrive/My Drive/Tran Lab Shared/Projects/Doris Duke PfSPZ Kenya/PfSPZ Nanostring Validation/NanoString PlexSet Design Tran_IU/Nanostring Validation Archive/Tran_IU_Hs_2_C3290-Summary.xlsx", range = "B3:M67", sheet = 1)

intersect(nanostring$`Customer Name`,myGenes)
```

### Load ExpressionSet

```{r load cpm eset, echo=TRUE, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- googledrive::drive_download(
  googledrive::as_id("123ihYx1ykdAvkMq7950yDLPSCTrSH089"), path = temp, overwrite = TRUE)
x <- readRDS(file = dl$local_path)
x$treat <- factor(x$treat, levels = c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
x$Timepoint <- as.integer(x$Timepoint)
```

### Reduce samples and genes

```{r reduce, echo=TRUE, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- googledrive::drive_download(
  googledrive::as_id("1k7olH6JtFd_hjRELj_HPmva7HZJqOEYt"), path = temp, overwrite = TRUE)
AllSPZSubjects <- readRDS(file = dl$local_path)
x <- x[,which(x$PATID %in% AllSPZSubjects)]
x <- x[,x$treat != "Placebo"]
x <- x[which(fData(x)$GeneSymbol %in% myGenes),]
rownames(x) <-  fData(x)$GeneSymbol
```

### Reshape, join, make survival table

```{r reshape and merge, echo=TRUE, message=FALSE, warning=FALSE}
cpm_lfc_mat <- pData(x) %>%
  mutate(Outcome = ifelse(mal.atp.3 == 1, "not protected", "protected")) %>%
  dplyr::select(SAMPLEID, PATID, Timepoint, Outcome) %>%
  left_join(., exprs(x) %>%
              t() %>%
              as.data.frame() %>%
              rownames_to_column(var = "SAMPLEID"),
            by = "SAMPLEID") %>%
  dplyr::select(-c(SAMPLEID)) %>%
  arrange(PATID, Timepoint) %>%
  pivot_longer(cols = 4:ncol(.), names_to = "GeneSymbol", values_to = "logCPM") %>%
  pivot_wider(names_from = "Timepoint", values_from = "logCPM") %>%
  mutate(logFC = `25`-`0`) %>%
  dplyr::select(PATID, GeneSymbol, logFC) %>%
  pivot_wider(names_from = GeneSymbol, values_from = logFC) %>%
  column_to_rownames(var = "PATID") %>%
  t() %>%
  as.data.frame()%>%
  dplyr::select(pData(x)[pData(x)$Timepoint==0,]$PATID)

x_delta <- x[,pData(x)$Timepoint==0]
rownames(pData(x_delta)) <- x_delta$PATID
Biobase::exprs(x_delta) <- as.matrix(cpm_lfc_mat)

survdat <- Biobase::exprs(x_delta) %>%
        as.data.frame() %>%
        rownames_to_column(var = "GeneSymbol") %>%
        dplyr::select(GeneSymbol, everything()) %>%
        pivot_longer(2:ncol(.), names_to = "PATID", values_to = "expression") %>%
        group_by(GeneSymbol) %>%
        mutate(exprs_up_down = factor(ifelse(expression>0, "induced", "not induced"), levels = c("not induced", "induced"))) %>%
        left_join(., fData(x_delta) %>%
                    dplyr::select(c(GeneSymbol, description)), by = "GeneSymbol") %>%
        left_join(., pData(x_delta) %>%
                    mutate(Outcome = ifelse(mal.atp.3 == 1, "not protected", "protected")) %>%
                    dplyr::select(c(SAMPLEID, PATID, Timepoint, treat, Outcome, site, SEX, age.vax1, mal.dvax, mal.dvax.tot, contains("atp"))), by = "PATID") %>%
        mutate(exprs_up_down_outcome = paste(Outcome, exprs_up_down, sep = " "))
```

### Kaplan-Meier plots

RP3-330M21.5 and ST3GAL3 are in Figure 4I of preprint

```{r km plots, echo=TRUE, message=FALSE, warning=FALSE}
myFollowup <- 6 #3 or 6
N <- length(unique(survdat$PATID))
survdat.filtered <- fit <- mySurvPlot <- c()
for(i in myGenes){
  survdat.filtered[[i]] <- survdat %>%
    filter(GeneSymbol == i)
  fit[[i]] <- survfit(Surv(tte.mal.atp.6, mal.atp.6) ~ exprs_up_down, data= survdat.filtered[[i]])
  mySurvPlot[[i]] <- ggsurvplot(fit[[i]], risk.table = TRUE, pval = TRUE,
                                break.time.by = 30, title = unique(survdat.filtered[[i]]$GeneSymbol),
                                xlab = "Days since vaccination",
                                ylab = "% free of parasitemia",
                                censor = TRUE,
                                palette = "lancet",
                                conf.int = TRUE,
                                conf.int.alpha = 0.1,
                                font.family = "Arial")
}
```


```{r print km plots, echo = FALSE, fig.align='center', fig.height = 5, fig.width = 6}
for(i in myGenes){
  plot.new()
  print(mySurvPlot[[i]], newpage = FALSE)
}
```

```{r print invidividual plots, echo=FALSE, include=FALSE}
tempdir <- "/Volumes/GoogleDrive/My Drive/Tran Lab Shared/Projects/Doris Duke PfSPZ Kenya/Tuan PfSPZ/KenyaPfSPZ/PfSPZ_R_Figures/"
cairo_pdf(filename = paste0(tempdir,"KSPZV1 Fig 4I Delta KM ST3GAL3.pdf"), width = 5.5, height = 5.5)
print(mySurvPlot$ST3GAL3)
dev.off()

cairo_pdf(filename = paste0(tempdir,"KSPZV1 Fig 4I Delta KM RP3.pdf"), width = 5.5, height = 5.5)
print(mySurvPlot$`RP3-330M21.5`)
dev.off()
```

### Cox Regression

```{r do cox regression to separate dose groups, echo=TRUE, message=FALSE, warning=FALSE}
cox.res <- c()
for(i in names(survdat.filtered)){
  survdat.cp <- survdat.filtered[[i]]
  survdat.cp$start		<-		0
  survdat.cp$stop		<-		survdat.cp$tte.mal.atp.6
  subj.pcrp <- rep(survdat.cp$start, as.vector(table(survdat.cp$SAMPLEID)))
  survdat.cp$start1 <- survdat.cp$start-subj.pcrp
  survdat.cp$stop1 <- survdat.cp$stop-subj.pcrp
  survdat.cp <- survdat.cp %>%
    dplyr::filter(stop1 > 0) %>%
    droplevels()
  fit <- coxph(Surv(start1, stop1, mal.atp.6) ~ exprs_up_down + age.vax1 + as.factor(site) + as.factor(mal.dvax) + as.factor(treat), data = survdat.cp)
  #vfit <- cox.zph(fit)
  #par(mfrow=c(3,2))
  #plot(vfit)
  #cox.zph(fit)
  #fit
  cox.res[[i]] <- as.data.frame(cbind(fit$n, fit$nevent, summary(fit)$conf.int[,c(1,3:4)], summary(fit)$coefficient[,5]))
  colnames(cox.res[[i]]) <- c("n","number of events","HR", "LCI", "UCI", "P value")
  cox.res[[i]] <- cox.res[[i]] %>%
    as.matrix() %>%
    signif(., digits = 3) %>%
    as.data.frame()
  cox.res[[i]]$Gene <- i
  rownames(cox.res[[i]]) <- c("upregulated post-vax (ref:downregulated post-vax)",
                              "age (months)",
                              "Wagai (ref:Siaya)",
                              "Pf infection during vax period (ref:no Pf during vax period)",
                              "9.0x10^5 PfSPZ (ref:4.5x10^5 PfSPZ)",
                              "1.8x10^6 PfSPZ (ref:4.5x10^5 PfSPZ)")
  cox.res[[i]] <- cox.res[[i]] %>%
    rownames_to_column(var = "Covariate") %>%
    dplyr::select(Gene, everything())
  }
  cox.res.bound <- bind_rows(cox.res, .id = "Gene")
```

### Table S4 of preprint

```{r show table, echo=FALSE}
  knitr::kable(cox.res.bound)
```


```{r write table to file, echo=FALSE, eval=FALSE}
resdir <- "/Volumes/GoogleDrive/My Drive/Tran Lab Shared/Projects/Doris Duke PfSPZ Kenya/Tuan PfSPZ/KenyaPfSPZ/PfSPZ_R_Results/Cox results"
writexl::write_xlsx(cox.res.bound, paste0(resdir, "Cox PH ", "for ", length(names(survdat.filtered)), " genes", " allVax postvax ", myFollowup, " mos_", Sys.Date(),".xlsx"))
```
