---
title: "all features model"
author: "Leetah Senkpeil"
date: "5/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, include=FALSE}
library(openxlsx)
library(xgboost)
library(dplyr)
library(tidyverse)
library(fgsea)
library(glmnet)
library(pROC)
library(aod)
library(caret)
library(limma)
library(caTools)
library(e1071)
library(robustbase)

datadir <- "D:\\Research\\Manuscript\\"
plotdir <- "D:\\Research\\Manuscript\\Baseline All Features\\"
#load("D:\\Research\\My Scripts\\PfSPZ Scripts\\Baseline ML parameters_Allfeats_two_thirds_train.RData")

```

Load Expression Data (logCPM) and calculate pathway activity scores of modules interest
```{r PAS}
#calculate pathway activity scores from log CPM expression values
#read in eset
pfspz_cpm_SM6 <- readRDS(paste0(datadir,"PfSPZ_cpm_ExpressionSet_244x21815_AllGroups_bothBatches_0_rmBatchFX_06082021_TMT.rds"))
lowBTMs <- readRDS(paste0(datadir,"lowBTMs.rds"))

# data <- readRDS(paste0(datadir,'Manuscript\\PfSPZ_cpm_ExpressionSet_244x21815_AllGroups_bothBatches_0_rmBatchFX_06082021_TMT.rds'))
# edata <- exprs(data)
# feat <- fData(data)
#myTimepoint <- "Baseline"#not necessary if using timepoint 0 eset only
myDoseGroup <- c("Placebo")

pfspz <- pfspz_cpm_SM6[,pfspz_cpm_SM6$treat == "Placebo"]

# #reduce expression set
# #pfspz_eset cmp.atp.3 has 0(no clinical malaria) and 1(clinical malaria) #change these to never_infected and infected
# pfspz_cpm_SM6$cmp.atp.3 <- gsub("0", "never_infected", pfspz_cpm_SM6$cmp.atp.3)
# pfspz_cpm_SM6$cmp.atp.3 <- gsub("1", "infected", pfspz_cpm_SM6$cmp.atp.3)
# 
# #pfspz_cpm_SM6 mal.atp.3 has 0(no parasitemia) and 1(parasitemia) #change these to never_infected and infected
# pfspz_cpm_SM6$mal.atp.3 <- gsub("0", "never_infected", pfspz_cpm_SM6$mal.atp.3)
# pfspz_cpm_SM6$mal.atp.3 <- gsub("1", "infected", pfspz_cpm_SM6$mal.atp.3)

#set options for reduction
#myBatch <- "Aug"
#myTimepoint <- "Baseline"
myDoseGroup <- c("1.8 x 10^6 PfSPZ")

pfspz <- pfspz_cpm_SM6[,pfspz_cpm_SM6$treat %in% myDoseGroup]

# #final reduction -- reduce exprs to just genes that are in the leading edge list
# madsno <- as.integer(nrow(pfspz)/5)
# mads <- apply(pfspz, 1, mad)  #mad filtering
# pfspz <- pfspz[ mads > sort(mads, decr=TRUE)[madsno],]
# #check size of eset
# dim(pfspz)

#scale data
a <- exprs(pfspz) <- t(scale(t(exprs(pfspz)), center = TRUE, scale = TRUE)) #generating z scores using scale function
rownames(a) <- fData(pfspz)$GeneSymbol
#take out genes in each module from complete expression data frame
annot.btm <- as.data.frame(unlist(names(lowBTMs)))
pas <- c()
for( i in 1:length(lowBTMs)){
  pas[[i]] <- colMedians(matrix(a[intersect(unique(gsub(" ", "",lowBTMs[[i]])),rownames(a)),], ncol = ncol(pfspz)), na.rm = TRUE)#pathway activity score defined as column means/medians depending how COLFUN is defined above 
  names(pas[[i]]) <- colnames(exprs(pfspz))
}
names(pas) <- names(lowBTMs)
#annot <- as.data.frame(paste(annot.btm[,1], annot.btm[,2]))
pas.df <- do.call("rbind", pas)
pas.df.annot <- merge(annot.btm, pas.df, by.x=1, by.y=0)
pas.df.annot <- na.omit(pas.df.annot)
rownames(pas.df.annot) <- pas.df.annot$`unlist(names(lowBTMs))`
pas.df.annot <- pas.df.annot[,-1]
```

Load Expression Data (logCPM) and calculate pathway activity scores of modules interest
```{r PAS}
#calculate pathway activity scores from log CPM expression values
#read in eset
# pfspz_cpm_SM6 <- readRDS(paste0(datadir,"PfSPZ_cpm_ExpressionSet_244x21815_AllGroups_bothBatches_0_rmBatchFX_06082021_TMT.rds"))
# lowBTMs <- readRDS(paste0(datadir,"lowBTMs.rds"))

#data <- readRDS(paste0(datadir,'Manuscript\\PfSPZ_cpm_ExpressionSet_244x21815_AllGroups_bothBatches_0_rmBatchFX_06082021_TMT.rds'))
# edata <- exprs(data)
# feat <- fData(data)
#myTimepoint <- "Baseline"#not necessary if using timepoint 0 eset only
# myDoseGroup <- c("Placebo")

plpfspz <- pfspz_cpm_SM6[,pfspz_cpm_SM6$treat == "Placebo"]

# #reduce expression set
# #pfspz_eset cmp.atp.3 has 0(no clinical malaria) and 1(clinical malaria) #change these to never_infected and infected
# pfspz_cpm_SM6$cmp.atp.3 <- gsub("0", "never_infected", pfspz_cpm_SM6$cmp.atp.3)
# pfspz_cpm_SM6$cmp.atp.3 <- gsub("1", "infected", pfspz_cpm_SM6$cmp.atp.3)
# 
# #pfspz_cpm_SM6 mal.atp.3 has 0(no parasitemia) and 1(parasitemia) #change these to never_infected and infected
# pfspz_cpm_SM6$mal.atp.3 <- gsub("0", "never_infected", pfspz_cpm_SM6$mal.atp.3)
# pfspz_cpm_SM6$mal.atp.3 <- gsub("1", "infected", pfspz_cpm_SM6$mal.atp.3)

#set options for reduction
#myBatch <- "Aug"
#myTimepoint <- "Baseline"


# #final reduction -- reduce exprs to just genes that are in the leading edge list
# madsno <- as.integer(nrow(pfspz)/5)
# mads <- apply(pfspz, 1, mad)  #mad filtering
# pfspz <- pfspz[ mads > sort(mads, decr=TRUE)[madsno],]
# #check size of eset
# dim(pfspz)

#scale data
pla <- exprs(plpfspz) <- t(scale(t(exprs(plpfspz)), center = TRUE, scale = TRUE)) #generating z scores using scale function
rownames(pla) <- fData(plpfspz)$GeneSymbol
#take out genes in each module from complete expression data frame
plannot.btm <- as.data.frame(unlist(names(lowBTMs)))
plpas <- c()
for( i in 1:length(lowBTMs)){
  plpas[[i]] <- colMedians(matrix(pla[intersect(unique(gsub(" ", "",lowBTMs[[i]])),rownames(pla)),], ncol = ncol(plpfspz)), na.rm = TRUE)#pathway activity score defined as column means/medians depending how COLFUN is defined above 
  names(plpas[[i]]) <- colnames(exprs(plpfspz))
}
names(plpas) <- names(lowBTMs)
#annot <- as.data.frame(paste(annot.btm[,1], annot.btm[,2]))
plpas.df <- do.call("rbind", plpas)
plpas.df.annot <- merge(plannot.btm, plpas.df, by.x=1, by.y=0)
plpas.df.annot <- na.omit(plpas.df.annot)
rownames(plpas.df.annot) <- plpas.df.annot$`unlist(names(lowBTMs))`
plpas.df.annot <- plpas.df.annot[,-1]
```

```{r select variables of interest}

pfspz_facs <- readRDS(paste0(datadir, "PfSPZ_FACS_PlasmaCytokine_MonocyteActivation_05012021 (1).rds"))

pfspz_facs <- pfspz_facs %>%
  #dplyr::select(!contains("Timepoint")) %>%
  dplyr::select(!contains("log2 fold-change anti-CSP IgG")) %>%
  dplyr::select(!contains("FACS_CD3+_of_live_lymphocytes")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_CD3+_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CD3+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_PfSPZ-specific_CD3+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_CD3+CD4+_of_live_lymphocytes")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_CD3+CD4+_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CD3+CD4+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_PfSPZ-specific_CD3+CD4+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_CD3+CD8+_of_live_lymphocytes")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_CD3+CD8+_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CD3+CD8+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_PfSPZ-specific_CD3+CD8+_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_CD19+_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CD19+_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_Plasmablasts_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_Plasmablasts_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_CSP-specific_memory_B_cells_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CSP-specific_memory_B_cells_of_live_leukocytes")) %>%
  # dplyr::select(!contains("FACS_CD11c+_of_live_monocytes")) %>%
  dplyr::select(!contains("FACS_CD11c+_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_CD14+_of_live_monocytes")) %>%
  dplyr::select(!contains("FACS_CD14+_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_CD56+CD16+_of_live_lymphocytes")) %>%
  dplyr::select(!contains("FACS_CD56+CD16+_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_memory_of_B_cells")) %>%
  dplyr::select(!contains("FACS_naive_of_B_cells")) %>%
  dplyr::select(!contains("FACS_IgG+_of_memory")) %>%
  dplyr::select(!contains("FACS_IgM+_of_memory")) %>%
  dplyr::select(!contains("FACS_IgG-IgM-_of_memory")) %>%
  dplyr::select(!contains("FACS_CSP-spec_of_memory_B_cells")) %>%
  dplyr::select(!contains("FACS_CSP-spec_of_B_cells")) %>%
  #dplyr::select(!contains("FACS_IgG_of_CSP-spec")) %>%
  #dplyr::select(!contains("FACS_IgM_of_CSP-spec")) %>%
  #dplyr::select(!contains("FACS_IgG-IgM-_of_CSP-spec")) %>%
  dplyr::select(!contains("FACS_PBs_of_memory")) %>%
  dplyr::select(!contains("FACS_PBs_of_B_cells")) %>%
  dplyr::select(!contains("FACS_atypical_memory_of_B_cells")) %>%
  dplyr::select(!contains("FACS_immature_of_B_cells")) %>%
  dplyr::select(!contains("FACS_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_TCRgd")) %>%
  dplyr::select(!contains("FACS_Vd1_of_T_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd1_of_T_cells")) %>%
  dplyr::select(!contains("FACS_Vd1_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd1_of_TCRgd")) %>%
  dplyr::select(!contains("FACS_Vd1/2-_of_T_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd1/2-_of_T_cells")) %>%
  dplyr::select(!contains("FACS_Vd1/2-_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd1/2-_of_TCRgd")) %>%
  dplyr::select(!contains("FACS_Vd2_of_T_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd2_of_T_cells")) %>%
  dplyr::select(!contains("FACS_Vd2_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vd2_of_TCRgd")) %>%
  dplyr::select(!contains("FACS_Vg9-_of_Vd2_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9-_of_Vd2_cells")) %>%
  dplyr::select(!contains("FACS_Vg9-Vd2+_of_T_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9-Vd2+_of_T_cells")) %>%
  dplyr::select(!contains("FACS_Vg9-Vd2+_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9-Vd2+_of_TCRgd")) %>%
  dplyr::select(!contains("FACS_Vg9+_of_Vd2_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9+_of_Vd2_cells")) %>%
  dplyr::select(!contains("FACS_Vg9+Vd2+_of_T_cells")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9+Vd2+_of_T_cells")) %>%
  dplyr::select(!contains("FACS_Vg9+Vd2+_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_PfSPZ-specific_Vg9+Vd2+_of_TCRgd")) %>%
  #dplyr::select(!contains("FACS_CD14+CD16+_of_live_monocytes")) %>%
  dplyr::select(!contains("FACS_CD14+CD16+_of_live_PBMCs")) %>%
  #dplyr::select(!contains("FACS_CD14+CD16-_of_live_monocytes")) %>%
  dplyr::select(!contains("FACS_CD14+CD16-_of_live_PBMCs")) %>%
  dplyr::select(!contains("FACS_CD14-CD16+_of_live_monocytes")) %>%
  # dplyr::select(!contains("CytokineObsConc_CCL2_MCP-1")) %>%
  # dplyr::select(!contains("CytokineObsConc_CCL4_MIP-1_beta")) %>%
  # dplyr::select(!contains("CytokineObsConc_CXCL10_IP-10")) %>%
  # dplyr::select(!contains("CytokineObsConc_GM-CSF")) %>%
  # dplyr::select(!contains("CytokineObsConc_IFN-alpha")) %>%
  # dplyr::select(!contains("CytokineObsConc_IFN-beta")) %>%
  # dplyr::select(!contains("CytokineObsConc_IFN-gamma")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-1_beta")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-10")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-12_p70")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-2")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-4")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-6")) %>%
  # dplyr::select(!contains("CytokineObsConc_IL-8_CXCL8")) %>%
  # dplyr::select(!contains("CytokineObsConc_TNF-alpha")) %>%
  dplyr::select(!contains("ICS_IL-1ß_pct_of_live_monocytes")) %>%
  dplyr::select(!contains("ICS_IL-1ß_Count")) %>%
  dplyr::select(!contains("ICS_IL-1ß_MFI_live_monocytes")) %>%
  dplyr::select(!contains("ICS_IL-8_pct_of_live_monocytes")) %>%
  dplyr::select(!contains("ICS_IL-8_Count")) %>%
  dplyr::select(!contains("ICS_IL-8_MFI_live_monocytes")) %>%
  dplyr::select(!contains("ICS_TNF_pct_of_live_monocytes")) %>%
  dplyr::select(!contains("ICS_TNF_count")) %>%
  dplyr::select(!contains("ICS_TNF_MFI_live_monocytes")) %>%
  dplyr::select(!contains("ICS_CD86_MFI_live_monocytes")) %>%
  dplyr::select(!contains("ICS_HLA-DR_MFI_live_monocytes"))
  #dplyr::select(!contains("FACS_CD14-CD16+_of_live_PBMCs")) ######this line was shifted down

```
make dataset with all features for feature elimination
```{r all features}


#order columns like pas.df.annot
pas.temp <- pas.df.annot
subj <- colnames(pas.temp)
ind <- which(pfspz_facs$PATID.OG %in% subj)
facs.temp <- as.data.frame(pfspz_facs[ind,])
rownames(facs.temp) <- facs.temp$PATID.OG
facs.temp <- t(facs.temp[,-c(1,2)])
facs.temp <- facs.temp[,order(colnames(facs.temp))]
pas.temp <- pas.temp[,order(colnames(pas.temp))]

pas.temp <- rbind(facs.temp,pas.temp)

#imputation of missing features
pas2 <- t(pas.temp)
inds <- which(rowSums(is.na(pas2)) > nrow(pas2)/2)

#lymph <- grep("live_lymphocytes", colnames(pas2))
#pbmcs <- grep("live_PBMCs", colnames(pas2))
#pf_sp <- grep("PfSPZ-specific",colnames(pas2))

library(missMDA)
#pas2 <- pas2[,lymph]
if(length(inds) > 0){
  pas_imp <- pas2[-inds,]
} else {pas_imp <- pas2}

cn <- colnames(pas_imp)
pas_imp <- t(pas_imp)
imp <- MIPCA(pas_imp)
pas_imputed <- imp$res.imputePCA

pas.df.annot <- pas_imputed

```

make dataset with all features for feature elimination
```{r all features}


#order columns like pas.df.annot
plpas.temp <- plpas.df.annot
plsubj <- colnames(plpas.temp)
plind <- which(pfspz_facs$PATID.OG %in% plsubj)
plfacs.temp <- as.data.frame(pfspz_facs[plind,])
rownames(plfacs.temp) <- plfacs.temp$PATID.OG
plfacs.temp <- t(plfacs.temp[,-c(1,2)])
plfacs.temp <- plfacs.temp[,order(colnames(plfacs.temp))]
plpas.temp <- plpas.temp[,order(colnames(plpas.temp))]

plpas.temp <- rbind(plfacs.temp,plpas.temp)

#imputation of missing features
plpas2 <- t(plpas.temp)
plinds <- which(rowSums(is.na(plpas2)) > nrow(plpas2)/2)

#lymph <- grep("live_lymphocytes", colnames(pas2))
#pbmcs <- grep("live_PBMCs", colnames(pas2))
#pf_sp <- grep("PfSPZ-specific",colnames(pas2))

library(missMDA)
#pas2 <- pas2[,lymph]
if(length(plinds) > 0){
  plpas_imp <- plpas2[-plinds,]
} else {plpas_imp <- plpas2}

plcn <- colnames(plpas_imp)
plpas_imp <- t(plpas_imp)
plimp <- MIPCA(plpas_imp)
plpas_imputed <- plimp$res.imputePCA

plpas.df.annot <- plpas_imputed

```

```{r recursive feature elimination}

model_df <- t(pas.df.annot)
#model_df <- model_df[-c(1,2),]
outcome <- as.data.frame(pfspz[,which(pfspz$SAMPLEID %in% rownames(model_df))]$mal.atp.3)
rownames(outcome) <- pfspz[,which(pfspz$SAMPLEID %in% rownames(model_df))]$SAMPLEID
colnames(outcome) <- "Outcome3mos"
outcome$Outcome3mos <-  gsub("never_infected", 0, outcome$Outcome3mos)
outcome$Outcome3mos <- gsub("infected", 1, outcome$Outcome3mos)

outcome <- as.factor(unlist(outcome))


set.seed(0623)
temp <- sample_n(as.data.frame(model_df), (nrow(model_df)*(2/3)))
indices <- which(rownames(model_df) %in% rownames(temp))


#check size of eset
dim(pfspz)
model_df_tr <- data.frame(model_df[indices,])
outcome_tr <- outcome[indices]

model_df_t <- data.frame(model_df[-indices,])
outcome_t <- outcome[-indices]

featlist <- list()
feat_seeds <- list()
for(iii in c(1:5)){
  temp.seed <- sample.int(10000,1)[[1]]
  set.seed(temp.seed)
  feats <- rfe(model_df_tr, outcome_tr, sizes = 3:5, metric = 'Accuracy', maximize = TRUE, rfeControl = rfeControl())
  featlist[[iii]] <- feats$optVariables[1:5]
  feat_seeds[[i]] <- temp.seed
}


feats
```

```{r find feats}

temp <- unlist(featlist[1])
#temp <- t(temp)


for(i in 2:5){
  temp2 <- unlist(featlist[i])
  temp <- c(temp,temp2)

}


feat_df <- as.data.frame(unique(temp))
for(i in 1:17){ #change to match num of unique features
  feat_df[i,2] <- length(which(temp == feat_df[i,1]))
}

feat_df <- feat_df[order(feat_df[,2], decreasing = T),]#order feats decreasing from most to least chosen


```

with selected features (PAS) use xgboost to predict infection status from transcriptional data
```{r machine learning}


# #feats <- feat_df[c(1:4),1]#set features to top 5 most common features from 100 rds feat. selection
feats <- c("FACS_CD14+_of_live_monocytes","spliceosome (M250)","proteasome (M226)", "MHC-TLR7-TLR8 cluster (M146)")
#feats <- c("FACS_CSP-specific_memory_B_cells_of_live_lymphocytes", "FACS_CD56+CD16+_of_live_lymphocytes", "CytokineObsConc_IL-8_CXCL8",
   #        "cytoskeleton/actin (SRF transcription targets) (M145.0)", "NK cell surface signature (S1)")
pas_df <- pas.df.annot

plpas_df <- plpas.df.annot
#pas_df$`Module title` <- paste(pas.df.annot$ID,pas.df.annot$`Module title`, sep = ": ")

common <- intersect(rownames(pas_df), feats)
plcommon <- intersect(rownames(plpas_df), feats)
model_df <- as.data.frame(t(pas_df[which(rownames(pas_df) %in% common),]))
plmodel_df <- as.data.frame(t(plpas_df[which(rownames(plpas_df) %in% plcommon),]))


pfspz_train <- pfspz_cpm_SM6[,pfspz_cpm_SM6$SAMPLEID %in% rownames(model_df)]
outcome <- as.data.frame(pfspz_train$mal.atp.3)
rownames(outcome) <- pfspz_train$SAMPLEID
colnames(outcome) <- "Outcome3mos"
outcome$Outcome3mos <- as.character(outcome$Outcome3mos)

pfspz_test <- pfspz_cpm_SM6[,pfspz_cpm_SM6$SAMPLEID %in% rownames(plmodel_df)]
ploutcome <- as.data.frame(pfspz_test$mal.atp.3)
rownames(ploutcome) <- pfspz_test$SAMPLEID
colnames(ploutcome) <- "Outcome3mos"



#check that outcome and model_df are in the same order
model_df$SubjID <- rownames(model_df)
outcome$SubjID <- rownames(outcome)
length(match(rownames(outcome), rownames(model_df)))
model_df <- left_join(model_df, outcome)
model_df$Outcome3mos <- gsub("never_infected", 0, model_df$Outcome3mos)
model_df$Outcome3mos <- gsub("infected", 1, model_df$Outcome3mos)
rownames(model_df) <- model_df$SubjID
model_df <- model_df[,-which(colnames(model_df) == "SubjID")] #Removing SubjID

#check that outcome and model_df are in the same order
plmodel_df$SubjID <- rownames(plmodel_df)
ploutcome$SubjID <- rownames(ploutcome)
length(match(rownames(ploutcome), rownames(plmodel_df)))
plmodel_df <- left_join(plmodel_df, ploutcome)
plmodel_df$Outcome3mos <- gsub( 0,"never_infected", plmodel_df$Outcome3mos)
plmodel_df$Outcome3mos <- gsub(1, "infected", plmodel_df$Outcome3mos)
plmodel_df$Outcome3mos <- gsub( "never_infected",1, plmodel_df$Outcome3mos)
plmodel_df$Outcome3mos <- gsub( "infected",0, plmodel_df$Outcome3mos)

rownames(plmodel_df) <- plmodel_df$SubjID
plmodel_df <- plmodel_df[,-which(colnames(plmodel_df) == "SubjID")] #Removing SubjID



#indices2 <- which(rownames(model_df) %in% rownames(model_df_tr))
save_indices <- indices

#create training and test sets
train_df <- model_df[indices,]
train.rows <- rownames(train_df)
test_df <- plmodel_df#[-indices,]
test.rows <- rownames(test_df)


train <- matrix(as.numeric(as.character(unlist(train_df))), ncol = ncol(train_df))
colnames(train) <- colnames(train_df)
test <- matrix(as.numeric(unlist(test_df)), ncol = ncol(test_df))
colnames(test) <-  colnames(test_df)
all <- matrix(as.numeric(unlist(model_df)), ncol = ncol(model_df))
colnames(all) <-  colnames(model_df)

#build xgboost datasets
d.train <- xgb.DMatrix(data = train[,-ncol(train)], label = (train[,ncol(train)]))#
d.test <- xgb.DMatrix(data = test[,-ncol(test)], label = (test[,ncol(test)]))#indices need to match training
#d.all <- xgb.DMatrix(data = all[,-ncol(all)], label = (all[,ncol(all)]))

temp.seed <- sample.int(10000,1)[[1]]

#Looping to train best parameters

best_param <- list()
best_seednumber = temp.seed
best_error = Inf

featlist <- list()
list_param <- list()
error_list <- list()

# for (i in 1:50) {
#   rm(model_df)
#   start_df <- t(pas.df.annot)
#   #model_df <- model_df[-c(1,2),]
#   outcome <- as.data.frame(pfspz[,which(pfspz$SAMPLEID %in% rownames(start_df))]$mal.atp.3)
#   rownames(outcome) <- pfspz[,which(pfspz$SAMPLEID %in% rownames(start_df))]$SAMPLEID
#   colnames(outcome) <- "Outcome3mos"
#   outcome$Outcome3mos <-  gsub("never_infected", 0, outcome$Outcome3mos)
#   outcome$Outcome3mos <- gsub("infected", 1, outcome$Outcome3mos)
#   
#   outcome <- as.factor(unlist(outcome))
#   
#   
#   temp <- sample_n(as.data.frame(start_df), (nrow(start_df)*(2/3)))
#   indices <- which(rownames(start_df) %in% rownames(temp))
#   
#   
#   #check size of eset
#   dim(pfspz)
#   model_df_tr <- data.frame(start_df[indices,])
#   outcome_tr <- outcome[indices]
#   
#   model_df_t <- data.frame(start_df[-indices,])
#   outcome_t <- outcome[-indices]
#   
#   feats <- rfe(model_df_tr, outcome_tr, sizes = 3:5, metric = 'Accuracy', maximize = TRUE, rfeControl = rfeControl())
#   
#   best_param <- list()
#   best_seednumber = temp.seed
#   best_error = Inf
#   
#   pas_df <- pas.df.annot
#   #pas_df$`Module title` <- paste(pas.df.annot$ID,pas.df.annot$`Module title`, sep = ": ")
#   
#   
#   if(length(feats$optVariables) >= 4){
#   model_ind <- c(grep(feats$optVariables[1], rownames(pas_df)),grep(feats$optVariables[2], rownames(pas_df)),grep(feats$optVariables[3], rownames(pas_df)),grep(feats$optVariables[4], rownames(pas_df)))}
#   else if(length(feats$optVariables) == 3){
#     odel_ind <- c(grep(feats$optVariables[1], rownames(pas_df)),grep(feats$optVariables[2], rownames(pas_df)),grep(feats$optVariables[3],rownames(pas_df)))
#   }
#   
  
  # model_df <- as.data.frame(t(pas_df[model_ind,]))
  # 
  # pfspz_train <- pfspz_cpm_SM6[,pfspz_cpm_SM6$SAMPLEID %in% rownames(model_df)]
  # outcome <- as.data.frame(pfspz_train$mal.atp.3)
  # rownames(outcome) <- pfspz_train$SAMPLEID
  # colnames(outcome) <- "Outcome3mos"
  # 
  # 
  # #check that outcome and model_df are in the same order
  # model_df$SubjID <- rownames(model_df)
  # outcome$SubjID <- rownames(outcome)
  # length(match(rownames(outcome), rownames(model_df)))
  # model_df <- left_join(model_df, outcome)
  # model_df$Outcome3mos <- gsub("never_infected", 0, model_df$Outcome3mos)
  # model_df$Outcome3mos <- gsub("infected", 1, model_df$Outcome3mos)
  # rownames(model_df) <- model_df$SubjID
  # model_df <- model_df[,-which(colnames(model_df) == "SubjID")] #Removing SubjID
  # 
  # #indices2 <- which(rownames(model_df) %in% rownames(model_df_tr))
  # save_indices <- indices
  # 
  # #create training and test sets
  # train_df <- model_df[indices,]
  # train.rows <- rownames(train_df)
  # test_df <- model_df[-indices,]
  # test.rows <- rownames(test_df)
  # 
  # 
  # train <- matrix(as.numeric(as.character(unlist(train_df))), ncol = ncol(train_df))
  # colnames(train) <- colnames(train_df)
  # test <- matrix(as.numeric(unlist(test_df)), ncol = ncol(test_df))
  # colnames(test) <-  colnames(test_df)
  # all <- matrix(as.numeric(unlist(model_df)), ncol = ncol(model_df))
  # colnames(all) <-  colnames(model_df)
  # 
  # #build xgboost datasets
  # d.train <- xgb.DMatrix(data = train[,-ncol(train)], label = (train[,ncol(train)]))#
  # d.test <- xgb.DMatrix(data = test[,-ncol(test)], label = (test[,ncol(test)]))#indices need to match training
  # #d.all <- xgb.DMatrix(data = all[,-ncol(all)], label = (all[,ncol(all)]))
  
  
  for (iter in 1:1000) {
    print(glue::glue("Curr Iter: ", iter, " Best Error: ", best_error))
    param <- list(objective = 'binary:logistic',
                  max_depth = sample(2:15,1),
                  eta = runif(1, 0.01, 0.5),
                  gamma = sample(c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10), 1),
                  subsample = runif(1, 0, 1),
                  colsample_bytree = runif(1, 0, 1),
                  min_child_weight = sample(0:10, 1),
                  lambda = sample(c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10), 1),
                  labmda_bias = sample(c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10), 1),
                  alpha = sample(c(0, 0.1, 0.2, 0.5, 1, 2, 5, 10), 1)
    )
    
    
    
    cv.nround = 1000
    cv.nfold = 3
    seed.number = sample.int(10000,1)[[1]]
    set.seed(seed.number)
    mdcv <- xgb.cv(data=d.train, verbose = F, params = param, nfold = cv.nfold,
                      nrounds = cv.nround, early_stopping_rounds = 30, maximize = FALSE, metrics = "error")
    min_error = min(mdcv$evaluation_log$test_error_mean)
  
    if (min_error < best_error) {
      best_error = min_error
      best_nround = which.min(mdcv$evaluation_log$test_error_mean)
      best_seednumber = seed.number
      best_param = param
      
     
      
    }
     
  }
#}


#best_param2 <- best_param #from saved Rdata file
# best_param2$gamma <- 1
# best_param2$lambda <- 0.75
# best_param2$eta <- .2880929
# best_param2$alpha <- 0.1
# best_param2$max_depth <- 5
# best_param2$colsample_bytree <- 0.269194
# best_param2$min_child_weight <- 0

temp.seed <- sample.int(10000,1)[[1]]
set.seed(3549)#set from saved data
model <- xgb.train(data = d.train, params = best_param2,nrounds = 50000, watchlist = list(val = d.test, train = d.train), early_stopping_rounds = 400, print_every_n = 10, maximize = F, eval_metric = "error")

# 3131
#3549
#9647
#2443
#3497
#5863
# seed1 <- temp.seed #23% val error
# seed2 <- temp.seed #23% val error
# seed3 <- temp.seed #14% val error
# seed4 <- temp.seed #19% val error
# seed5 <- temp.seed #23% val error
# seed6 <- temp.seed #19% val error
# seed7 <- temp.seed #23% val error
# seed8 <- temp.seed #28% val error
# seed9 <- temp.seed #19% val error
# seed10 <- temp.seed #33% val error
# seed11 <- temp.seed #23% val error
# seed12 <- temp.seed #28% val error
# models with placebo feature selection features
# seed1a <- temp.seed #23.8% val error, 232
# seed2a <- temp.seed #23% val error
# seed3a <- temp.seed #23% val error
# seed4a <- temp.seed #23% val error
# seed5a <- temp.seed #23% val error
# seed6a <- temp.seed #19% val error
# seed7a <- temp.seed #33% val error
# seed8a <- temp.seed #28% val error
# seed9a <- temp.seed #19% val error
# seed10a <- temp.seed #33% val error
# seed11a <- temp.seed #28% val error
# seed12a <- temp.seed #19% val error

pred <- predict(model, newdata = test[,c(1:(ncol(test)-1))])

(sum((pred >= 0.5) == (test[,ncol(test)] == 1)))/length(test[,ncol(test)])

imp <- xgb.importance(feature_names = colnames(train[,-5]), model = model) 
xgb.plot.importance(importance_matrix = imp)

plot_imp <- ggplot(data = imp, aes(x = Importance, y = reorder(Feature, Importance))) +
    geom_bar(stat = 'identity',fill = "#fef0d9", width = 0.95) +
    theme_classic(base_family = "sans", base_size = 6) +
    theme(axis.ticks = element_line(color = "grey80")) +
    ggtitle("Baseline Transcriptomics Feature Importance") +
    xlab("Importance") + ylab("Features") +
    theme(plot.title=element_text(size =6, hjust = 1, vjust = -1), aspect.ratio = 0.75, axis.text.x.bottom =element_text(size =6)) +
    theme(legend.position = "none")
  
cairo_pdf(paste0(plotdir,"Placebo All Features baseline ML Importance Plot.pdf"),width = 3, height = 2.5)

print(plot_imp)

dev.off()

```

The purpose of this chunk is to make use of xgboostExplainer to understand the impact of each feature on the model -- making logistic regression model more transparent
```{r xgb explain}
library(xgboostExplainer)

#predictions
pred.test <- predict(model, newdata = test[,c(1:(ncol(test)-1))])#using test data as newdata -- can use all data according to xgboost documentation
nodes.test <- predict(model, newdata = test[,c(1:(ncol(test)-1))], predleaf = TRUE)

#build tree using model nodes and features
trees <- xgb.model.dt.tree(colnames(d.train), model = model)

#step 1: build explainer
explainer <- buildExplainer(model, d.train, type = "binary")#build explanations for nodes/leafs
pred.breakdown <- explainPredictions(model, explainer, d.test)#predictions explained

#show waterfall
for (i in 1:nrow(test)){
water <- showWaterfall(model, explainer, d.test, test[,c(1:(ncol(test)-1))],i,type = "binary")

cairo_pdf(paste0(plotdir,"Waterfall\\All Features Baseline Waterfall_", seed2, "_", i, "_.pdf"),width = 3, height = 2.5)

print(water)

dev.off()  
}

trees

```

```{r xgb shap summary plot}
xgb.ggplot.shap.summary <- function(data, shap_contrib = NULL, features = NULL, top_n = 10, model = NULL,
                                    trees = NULL, target_class = NULL, approxcontrib = FALSE, subsample = NULL) {
  data_list <- xgboost:::xgb.shap.data(
    data = data,
    shap_contrib = shap_contrib,
    features = features,
    top_n = top_n,
    model = model,
    trees = trees,
    target_class = target_class,
    approxcontrib = approxcontrib,
    subsample = subsample,
    max_observations = 10000  # 10,000 samples per feature.
  )
  plot_outcome <- rep(test[,ncol(test)], 3)
  p_data <- xgboost:::prepare.ggplot.shap.data(data_list, normalize = TRUE)
  # Reverse factor levels so that the first level is at the top of the plot
  p_data[, "feature" := factor(feature, rev(levels(feature)))]
  p <- ggplot2::ggplot(p_data, ggplot2::aes(x = feature, y = p_data$shap_value, colour = p_data$feature_value)) +
    ggplot2::geom_jitter(alpha = 1, width = 0.1, height = 0.1, size = 2) +
    ggplot2::scale_colour_distiller(limits = c(-1.5, 1.5), na.value = 0, type = "div", palette = "Spectral", direction = 1) +
    ggplot2::geom_abline(slope = 0, intercept = 0, colour = "darkgrey") +
    ggplot2::coord_flip()

  p
}


```


ROC curves for ML model
```{r ggROC}
library(pROC)

roc1 <- pROC::roc(test[,ncol(test)], pred)
roc1 <- smooth(roc1)


ggroc <- function(roc, showAUC = TRUE, interval = 0.2, breaks = seq(0, 1, interval)){
  require(pROC)
  if(class(roc) != "roc")
    simpleError("Please provide roc object from pROC package")
  plotx <- rev(roc$specificities)
  ploty <- rev(roc$sensitivities)
  
  ggplot(NULL, aes(x = plotx, y = ploty)) +
    geom_segment(aes(x = 0, y = 1, xend = 1,yend = 0), alpha = 0.5, colour = "red") + 
    geom_step() +
    scale_x_reverse(name = "Specificity",limits = c(1,0), breaks = breaks, expand = c(0.001,0.001)) + 
    scale_y_continuous(name = "Sensitivity", limits = c(0,1), breaks = breaks, expand = c(0.001, 0.001)) +
    theme_bw(base_family = "sans", base_size = 6) +
    theme(axis.ticks = element_line(color = "grey80")) +
    coord_equal() + 
    annotate("text", x = interval*3, y = interval*3, vjust = 0, label = paste("AUC =",sprintf("%.3f",roc$auc)), family = "sans", size = 4) +
    geom_area(fill = "black", alpha = .4) +
    ggtitle("Baseline All Features Model")+
    theme(plot.title=element_text(size =4))
}

p <- ggroc(roc1)
p + scale_fill_hue(l = 45)

cairo_pdf(paste0(plotdir,"Placebo All feats baseline ML ROC.pdf"), width = 1.5, height = 1.5)

print(p)

dev.off()
# Set our cutoff threshold
pred.resp <- ifelse(pred >= 0.50, 1, 0)
 
# Create the confusion matrix
cm <- ModelMetrics::confusionMatrix(test[,ncol(test)],pred.resp)

cm

#create shap plot to visualize feature size and contribution to the model
shap <- xgb.ggplot.shap.summary(data = test[,c(1:(ncol(test)-1))], model = model) +
    theme_classic(base_family = "sans", base_size = 6) +
    theme(axis.ticks = element_line(color = "grey80")) +
    ggtitle("Baseline Global Feature Impact") +
    ylab("SHAP value (impact on model output)") + xlab(NULL) +
    theme(plot.title=element_text(size =6, hjust = 1, vjust = -1), aspect.ratio = .5, axis.text.x =element_text(size =6), axis.text.y =element_text(size =6)) +
    theme(legend.position = "bottom") + labs(color ="Feature Value")
  
cairo_pdf(paste0(plotdir,"Placebo Baseline SHapley Additive exPlanations All feats.pdf"),width = 3.5, height = 2)

print(shap)

dev.off()

39/64

```

```{r random forest}
install.packages("caTools")       # For sampling the dataset
install.packages("randomForest")  # For implementing random forest algorithm
  
# Loading package
library(caTools)
library(randomForest)
  
# Splitting data in train and test data
split <- sample.split(t(model_df), SplitRatio = 0.7)
split
  
train <- subset(model_df, split == "TRUE")
test <- subset(model_df, split == "FALSE")

train <- na.omit(train)
train$Outcome3mos <- as.numeric(train$Outcome3mos)
test <- na.omit(test)
  
# Fitting Random Forest to the train dataset
set.seed(120)  # Setting seed
classifier_RF = randomForest(x = train[-5],
                             y = train$Outcome3mos,
                             ntree = 500)
  
classifier_RF
  
# Predicting the Test set results
y_pred = predict(classifier_RF, newdata = test[-5])
pred.resp <- ifelse(y_pred >= 0.50, 1, 0)  
# Confusion Matrix
confusion_mtx = table(test[, 5], y_pred)
confusion_mtx

y_pred <- as.numeric(y_pred)
cm <- ModelMetrics::confusionMatrix(as.numeric(test[,ncol(test)]),pred.resp)

roc1 <- pROC::roc(as.numeric(test_set[,ncol(test_set)]), as.numeric(y_pred))
roc1 <- smooth(as.numeric(roc1))

caret::RMSE(as.numeric(test_set[,ncol(test_set)]),as.numeric(y_pred)) 
# Plotting model
plot(classifier_RF)
  
# Importance plot
importance(classifier_RF)
  
# Variable importance plot
varImpPlot(classifier_RF)
```

