---
title: "KSPZV1 PCA-LDA Delta Post-vax Plots"
author: "Aditi Upadhye"
date: "5/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(MASS)
library(googledrive)
library(ggord)
library(pcaExplorer)
figdir <- "/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Figures/Revised Figure Panels/"
```

## Objective

PCA-LDA for delta post-vax

```{r download file, echo=TRUE, message=FALSE, warning=FALSE}
  temp <- tempfile(fileext = ".rds")
  dl <- drive_download(
  as_id("17xx0YaggLiyWdJc9rqA1nkPSE7dj05p0"), path = temp, overwrite = TRUE)
   eset <- readRDS(dl$local_path)
```

## Reduce by groups

Reduce if needed, clean, wrangle

```{r reduce groups, echo=TRUE, message=FALSE, warning=FALSE}
myGroups <- c("Placebo","4.5 x 10^5 PfSPZ","9.0 x 10^5 PfSPZ","1.8 x 10^6 PfSPZ")
eset <- eset[,eset$treat %in% myGroups ]
eset$Outcome <- factor(ifelse(eset$mal.atp.3 == 0 , "P","NP"), levels = c("P","NP"))
eset$Class <- paste0(eset$treat,"_",eset$Outcome)
fData(eset) <- fData(eset) %>%
  as.data.frame() %>%
  remove_rownames()
rownames(fData(eset)) <- make.names(fData(eset)$GeneSymbol, unique = TRUE)
rownames(eset) <- rownames(fData(eset))
colnames(eset) <- pData(eset)$PATID
```

## PCA

### Scree plot

```{r prepare PCA and scree, echo=TRUE, fig.align='center', message=FALSE, warning=FALSE}
pca <- prcomp(t(exprs(eset)),
              center = TRUE,
              scale. = TRUE, rank. = 120)

## plot scree plot to determine proportion variation of principle components
proportion.variance <- summary(pca)
scree_plot <- pcaExplorer::pcascree(pca,pc_nr = 120, type = "cev")
scree_plot
```

### PCA Plot

```{r preprare data for plot, echo=TRUE, fig.align='center'}
pheno <- pData(eset)
pheno <- pheno %>%
  droplevels()
```

```{r option adding non-transcriptomic features, echo = FALSE, eval=FALSE}
pheno$SEX <- ifelse(pheno$SEX == "F", 1, 0)
pheno$site <- ifelse(pheno$site == "Siaya", 1, 0)
pheno[is.na(pheno$mal.dvax.tot),]$mal.dvax.tot <- median(pheno$mal.dvax.tot, na.rm = TRUE) #impute one missing value with median
pheno[is.na(pheno$pfcsp_pre),]$pfcsp_pre <- median(pheno$pfcsp_pre, na.rm = TRUE) #impute one missing value with median
#cbind mal.dvax.total to pca$x
if(all(rownames(pheno) == rownames(pca$x))){
  pca$x <- cbind(as.matrix(pheno[c("mal.dvax.tot","pfcsp_pre", "age.vax1", "SEX", "site")]), pca$x)
}
```

```{r model lda, echo=TRUE, fig.align='center', message=FALSE, warning=FALSE}
model_lda <- lda(pca$x, grouping = pheno$Class)
prop.lda = model_lda$svd^2/sum(model_lda$svd^2)
plda <- predict(object = model_lda)
dataset = data.frame(PATID = pheno[,"PATID"],
                     Class = pheno[,"Class"],
                     site = pheno[,"site"],
                     treat = pheno[,"treat"],
                     Age = pheno[,"age.vax1"],
                     Outcome = pheno[,"Outcome"],
                     Sex = pheno[,"SEX"],
                     Seqbatch = pheno[,"SEQBATCH"],
                     pfcsp_pre = pheno[,"pfcsp_pre"],
                     month.vax1 = pheno[,"month.vax1"],
                     mal.dvax = pheno[,"mal.dvax.tot"],
                     pca = pca$x,
                     lda = plda$x)

# PCA plot
prop.pca = pca$sdev^2/sum(pca$sdev^2)
pca_plot <- ggplot(dataset) + geom_point(aes(pca.PC1, pca.PC2, colour = treat, shape = Outcome), size = 2.5) +
  labs(x = paste("PC1 (", scales::percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", scales::percent(prop.pca[2]), ")", sep=""))+
  scale_color_manual(values=c("#808080","#fdcc8a","#fc8d59","#d7301f"))+
  scale_fill_manual(values=c("#808080","#fdcc8a","#fc8d59","#d7301f")) +
  theme_bw()
pca_plot
```

## PCA-LDA

Add LDA vectors to plot as separate geom. 


```{r lda and top ld, echo =FALSE}
# obs <- plda$x
# vars <- model_lda$scaling
# 
# whichpc <- "LD1"
# topN <- 1
# sorted_LD1 <- sort(vars[, whichpc]) 
# extreme_LD1 <- c(tail(sorted_LD1, topN), head(sorted_LD1, topN)) 
# extreme_LD1 <- extreme_LD1 %>%
#   as.data.frame() %>%
#   rownames_to_column("PC")%>%
#   rename("LD1"=".")
# 
# whichpc <- "LD2"
# topN <- 1
# sorted_LD2 <- sort(vars[, whichpc]) 
# extreme_LD2 <- c(tail(sorted_LD2, topN), head(sorted_LD2, topN)) 
# extreme_LD2 <- extreme_LD2 %>%
#   as.data.frame() %>%
#   rownames_to_column("PC")%>%
#   rename("LD2"=".")
# 
# my_loadings <- c(extreme_LD1$PC,extreme_LD2$PC)
```

```{r lda and top ld hypotenuse, warning=FALSE, message=FALSE}
obs <- plda$x
vars <- model_lda$scaling
vars_hypotenuse <- vars[,c("LD1","LD2")] %>%
  as.data.frame() %>%
  rownames_to_column("PC") %>%
  mutate(hypotenuse = sqrt(LD1^2+LD2^2)) %>%
  mutate(direction = ifelse(LD1>0 & LD2>0, "pospos",
                            ifelse(LD1>0 & LD2<0, "posneg",
                                   ifelse(LD1<0 & LD2<0, "negneg","negpos")))) %>%
  group_by(direction) %>%
  dplyr::arrange(desc(hypotenuse)) %>%
  slice(1) %>%
  ungroup()
  
my_loadings <- vars_hypotenuse$PC
```

### Table of PCs that discriminate classes the most by quadrant (PC with max hypotenuse in each quadrant)

```{r table of hypo, echo=FALSE}
knitr::kable(vars_hypotenuse)
```


```{r add LDA, echo = TRUE, fig.align='center', message=FALSE, warning=FALSE}
dat <- dataset %>% 
  mutate(
    one = plda$x[, 'LD1'], 
    two = plda$x[, 'LD2'], 
    newgrp = sample(c('a', 'b'), n(), replace = T)
    )
# dat2 <- dataset %>% 
#   mutate(
#     one = plda$x[, 'LD2'], 
#     two = plda$x[, 'LD3'], 
#     newgrp = sample(c('a', 'b'), n(), replace = T)
#     )
# dat3 <- dataset %>% 
#   mutate(
#     one = plda$x[, 'LD1'], 
#     two = plda$x[, 'LD3'], 
#     newgrp = sample(c('a', 'b'), n(), replace = T)
    # )
dat$treat <- factor(dat$treat,levels = c("Placebo","4.5 x 10^5 PfSPZ","9.0 x 10^5 PfSPZ","1.8 x 10^6 PfSPZ"))
dat$Class <- factor(dat$Class,levels = c("Placebo_NP","Placebo_P","4.5 x 10^5 PfSPZ_NP","4.5 x 10^5 PfSPZ_P",
 "9.0 x 10^5 PfSPZ_NP","9.0 x 10^5 PfSPZ_P","1.8 x 10^6 PfSPZ_NP","1.8 x 10^6 PfSPZ_P"))
```

## Figure 4A of revised manuscript

```{r biplot with top/bottom principle components, echo =TRUE, message=FALSE, warning=FALSE}

x <- "LD1"
y <- "LD2"

data <- data.frame(obsnames=row.names(plda$x), plda$x)
datapc <- data.frame(varnames=rownames(model_lda$scaling), model_lda$scaling)
datapc <- datapc %>%
              dplyr::select(varnames,LD1,LD2)%>%
              filter(varnames %in% my_loadings)

datapc <- transform(datapc,
            v1 = 55 * get(x),
            v2 = 55 * get(y)
)
    
biplot <- ggord(model_lda, dat$Class,axes = c("1", "2"), obslab = FALSE, ptslab = FALSE, ellipse = TRUE, parse = FALSE,
                  vec_lab = (''),veccol = 'transparent')
biplot$layers[[1]] <- NULL
biplot <- biplot +
  geom_hline(yintercept = 0, color = "gray40") +
  geom_vline(xintercept = 0, color = "gray40") +
  geom_point(data = dat, aes(shape = Outcome, color = Class, fill = Class), size = 4)+
  geom_segment(data=datapc, aes(x=0, y=0, xend=v1, yend=v2), arrow=arrow(length=unit(0.7,"cm")), alpha=1, color="black")+
  geom_text(data=datapc, aes(x=v1, y=v2, label=varnames), size = 6, vjust=1,color="black",position=position_jitter(width=1,height=1))+
  scale_shape_manual(values = c(21, 24)) + #21 (circle, P), #24 (triangle, NP)
  scale_color_manual(values=c(rep("grey85",2), rep("black",6))) + #outline of shapes
  scale_fill_manual(values=c("#6e6e6e","#6e6e6e","#fdcc8a","#fdcc8a","#fc8d59","#fc8d59","#d7301f","#d7301f")) + #shape color
  guides(fill = guide_legend(override.aes= list(shape = NA))) +
  theme(legend.title = element_blank(),
        axis.text = element_text(size = 14, family = "Arial"),
        axis.title = element_text(size = 16, family = "Arial"))

biplot
```

```{r save biplot, eval=FALSE, echo=FALSE}
cairo_pdf(paste0(figdir,"Fig 4A PCA-LDA biplot delta revision.pdf"), height = 6, width = 7)
biplot
dev.off()
```


```{r interpretation, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
pheno %>%
  as_tibble() %>%
  group_by(Class) %>%
  summarize(n = n(), mean_mal = mean(mal.dvax.tot, na.rm =TRUE), median_mal = mean(mal.dvax.tot, na.rm =TRUE)) %>%
  arrange(mean_mal)
``` 

```{r interpretation 2, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
pheno %>%
  as_tibble() %>%
  group_by(site) %>%
  summarize(n = n(), mean_mal = mean(mal.dvax.tot, na.rm =TRUE), median_mal = mean(mal.dvax.tot, na.rm =TRUE)) %>%
  arrange(mean_mal)

pheno %>%
  as_tibble() %>%
  group_by(mal.vax.1) %>%
  summarize(n = n(), mean_mal = mean(mal.dvax.tot, na.rm =TRUE), median_mal = mean(mal.dvax.tot, na.rm =TRUE)) %>%
  arrange(mean_mal)



table(pheno$Class, pheno$site)
table(pheno$mal.atp.3, pheno$site)
table(pheno$Class, pheno$site)
``` 


```{r apply gsea, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
library(devtools)
source_url("https://raw.githubusercontent.com/TranLab/ModuleLists/main/NamedGeneRankList2GseaTable.R")
resdir <- "/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Supplementary Tables/PC LDA GSEA tables/"

#PC86
ranks <- pca$rotation %>%
  as.data.frame() %>%
  rownames_to_column(var = "Gene") %>%
  as_tibble() %>%
  dplyr::select(c(Gene, PC86)) %>%
  arrange(desc(PC86)) %>%
  deframe()
PC86_GSEA <- NamedGeneRankList2GseaTable(rankedgenes = ranks, geneset = "bloodmodules", output_directory = resdir, filename_prefix = "PC86", minSize = 20, fixed_seed = TRUE)
PC86_GSEA <- PC86_GSEA %>%
  arrange(padj) %>%
  mutate("-log10 BH-adj p value" = log10(padj)) %>%
  mutate(PC = "PC86")

#PC93
ranks <- pca$rotation %>%
  as.data.frame() %>%
  rownames_to_column(var = "Gene") %>%
  as_tibble() %>%
  dplyr::select(c(Gene, PC93)) %>%
  arrange(desc(PC93)) %>%
  deframe()
PC93_GSEA <- NamedGeneRankList2GseaTable(rankedgenes = ranks, geneset = "bloodmodules", output_directory = resdir, filename_prefix = "PC93", minSize = 20, fixed_seed = TRUE)
PC93_GSEA <- PC93_GSEA %>%
  arrange(padj) %>%
  mutate("-log10 BH-adj p value" = log10(padj)) %>%
  mutate(PC = "PC93")

#PC97
ranks <- pca$rotation %>%
  as.data.frame() %>% #convert matrix to dataframe to allow subsequent data wrangling
  rownames_to_column(var = "Gene") %>%
  as_tibble() %>%
  dplyr::select(c(Gene, PC97)) %>% #select only PC of interest
  arrange(desc(PC97)) %>% #arrange in descending order
  deframe() #deframe to create named vector
PC97_GSEA <- NamedGeneRankList2GseaTable(rankedgenes = ranks, geneset = "bloodmodules", output_directory = resdir, filename_prefix = "PC97", minSize = 20, fixed_seed = TRUE)
PC97_GSEA <- PC97_GSEA %>%
  arrange(padj) %>%
  mutate("-log10 BH-adj p value" = log10(padj)) %>%
  mutate(PC = "PC97")

#PC108
ranks <- pca$rotation %>%
  as.data.frame() %>%
  rownames_to_column(var = "Gene") %>%
  as_tibble() %>%
  dplyr::select(c(Gene, PC108)) %>%
  arrange(desc(PC108)) %>%
  deframe()
PC108_GSEA <- NamedGeneRankList2GseaTable(rankedgenes = ranks, geneset = "bloodmodules", output_directory = resdir, filename_prefix = "PC108", minSize = 20, fixed_seed = TRUE)
PC108_GSEA <- PC108_GSEA %>%
  arrange(padj) %>%
  mutate("-log10 BH-adj p value" = log10(padj)) %>%
  mutate(PC = "PC108")
```

```{r bind and make heatmap, echo = TRUE, fig.align='center'}
bind_PC_GSEA <- bind_rows(PC86_GSEA, PC93_GSEA, PC97_GSEA, PC108_GSEA) %>%
  arrange(desc(pathway))

# bind_PC_GSEA_wide <- bind_PC_GSEA %>%
#   dplyr::select(c(pathway, PC, NES)) %>%
#   pivot_wider(.,names_from = PC, values_from = NES)
```

## Make Heatmap (Figure S5A of revised manuscript)

Only pathways/modules from low-level BTMs, high-level BTMs, and Monaco modules with BH-adjusted p value < 0.20 shown.

```{r make heatmap, echo = TRUE, fig.align='center', fig.width=4.75, fig.height=5, message=FALSE, warning=FALSE}
library(hrbrthemes)
textcol <- "grey40"
pc_heatmap <- bind_PC_GSEA %>%
  filter(module_type != "BloodGen3Module") %>%
  filter(padj<0.20) %>%
  droplevels() %>%
  ggplot(., aes(x = PC, y = fct_reorder(pathway, abs(NES), .desc = FALSE))) +
  geom_tile(aes(fill = NES), color = "white", size = 0.2) +
  scale_fill_gradient2(low = "blue", mid = "white", high= "red", 
                       midpoint =  0, na.value = "red") +
  scale_x_discrete(expand = c(0,0)) +
  ylab("module/pathway") +
  xlab("") +
  theme_ipsum(base_family = "Arial", base_size = 10,) +
  theme(legend.position="right", legend.direction="vertical",
        legend.title=element_text(colour=textcol),
        legend.margin=margin(grid::unit(0, "cm")),
        legend.text=element_text(colour=textcol, size=7, face="bold"),
        legend.key.height=grid::unit(0.25, "cm"),
        legend.key.width=grid::unit(0.2, "cm"),
        axis.text.x=element_text(size=10, colour=textcol, angle = 45, vjust = 1, hjust=1),
        axis.text.y=element_text(vjust=0.2, colour=textcol),
        axis.title = element_blank(),
        axis.ticks=element_line(size=0.4),
        plot.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major = element_blank(),
        plot.margin=margin(0.7, 0.4, 0.1, 2.75, "cm")
      )
pc_heatmap
```


```{r save heatmap, echo=FALSE, eval=FALSE}
cairo_pdf(filename = paste0(figdir, "Fig S5A delta PC heatmap.pdf"), width = 4.75, height = 3.75)
pc_heatmap
dev.off()
```
