---
title: "KSPZV1 DGE Δ (postvax/baseline)"
author: "Tuan M. Tran"
date: "09/16/2021"
output: html_document
---

### Objective

Perform differential gene expression using edgeR to assess differences betwee protected and not protected infants in the KSPZV1 malaria vaccine trial.
This analysis specifically evaluates differences in post-vax with adjustment for baseline.


```{r libraries, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(edgeR)
library(googledrive)
library(tidyverse)
library(fgsea)
library(data.table)
library(EDASeq)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load ExpressionSet

```{r load SeqExpressionSet 2, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("17P7RwDaPCwDXcrD82N4fYd3vGOcueFEH"), path = temp, overwrite = TRUE)
x <- readRDS(file = dl$local_path)
x$treat <- factor(x$treat, levels = c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
```

### Set options

```{r set options, message=FALSE, warning=FALSE}
myGroups <- unique(x$treat)
```


```{r more reduction, echo=FALSE}
xgroup <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(xgroup)){
  xgroup[[i]] <- x[, which(x$treat == i)]
  xgroup[[i]] <- xgroup[[i]][,xgroup[[i]]$Timepoint == 0 | xgroup[[i]]$Timepoint == 25]
  xgroup[[i]]$treat <- droplevels(xgroup[[i]]$treat)
  print(i)
  print(table(Outcome = xgroup[[i]]$Outcome, Timepoint = xgroup[[i]]$Timepoint))
  print(table(Outcome = xgroup[[i]]$Outcome, Dosegroup = xgroup[[i]]$treat))
  print(dim(xgroup[[i]]))
}
```

## Build DGEList Object and make Expression Sets

### Build DGEList Object

```{r make DGEList object, message=FALSE, warning=FALSE}
#Define group by dose, timepoint and outcome
ygroup <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(ygroup)){
  ygroup[[i]]  <- DGEList(counts=counts(xgroup[[i]]), genes=fData(xgroup[[i]]), group= factor(paste(xgroup[[i]]$treat, xgroup[[i]]$Timepoint, xgroup[[i]]$Outcome, sep = "_")), remove.zeros=T)
  ygroup[[i]]$samples$PATID <- gsub("_.*", "", rownames(ygroup[[i]]$samples))
}
```

### Filter out low expression features

```{r low expression filters, message=FALSE, warning=FALSE}
for(i in names(ygroup)){
  keep <- filterByExpr(ygroup[[i]])
  ygroup[[i]] <- ygroup[[i]][keep, , keep.lib.sizes=FALSE]
  ygroup[[i]]$samples$lib.size <- colSums(ygroup[[i]]$counts)
  ygroup[[i]] <- calcNormFactors(ygroup[[i]])	#Normalization
}
```

##Remove unpaired samples and then arrange by patient and timepoint

This is necessary for paired analyses

```{r remove unpaired samples then arrange}
for(i in names(ygroup)){
  ygroup[[i]] <- ygroup[[i]][,!is.na(ygroup[[i]]$samples$group)]
  ygroup[[i]] <- ygroup[[i]][,duplicated(ygroup[[i]]$samples$PATID) | duplicated(ygroup[[i]]$samples$PATID, fromLast = TRUE)]
  ygroup[[i]] <- ygroup[[i]][,order(ygroup[[i]]$samples$PATID, ygroup[[i]]$samples$group)]
  print(i)
  print(ifelse(all(gsub('_.*', '', colnames(ygroup[[i]][,grepl("_0", colnames(ygroup[[i]]))])) ==
               gsub('_.*', '', colnames(ygroup[[i]][,grepl("_25", colnames(ygroup[[i]]))]))) &
           all(ygroup[[i]]$samples$Timepoint == rep(c(0,25),length(ygroup[[i]]$samples$Timepoint)/2)), 
         "Samples are paired. All is good.","Stop and check order of samples."))
}
```

## Remake ExpressionSet after sample and feature (gene) filtering steps.

```{r remake SeqExpressionSet after gene filtering and sample reduction, echo = FALSE}
for(i in names(ygroup)){
  myintersection <- intersect(rownames(pData(xgroup[[i]])), colnames(ygroup[[i]]$counts))
  xgroup[[i]] <- newSeqExpressionSet(counts = ygroup[[i]]$counts[,myintersection],
                           featureData = AnnotatedDataFrame(fData(xgroup[[i]])[which(rownames(fData(xgroup[[i]])) %in% rownames(ygroup[[i]]$counts)),]),
                           phenoData = AnnotatedDataFrame(pData(xgroup[[i]])[myintersection,]))
}
```

## Design Matrix - Compare Protected (0) vs Not Protected (1)

Create design matrix according to EdgeR user guide (Ch 3.5 Comparisons Both Between and Within Subjects)  
Note that "intergenic_adj" should only be used if you use seq data that HAS NOT been SeqMonk corrected. Choose no_adj when using SeqMonk corrected data.

### Analysis between Protected and Susceptible within a Dose Group

This chunk is the analysis used for final submission.

```{r, build design matrix afs protected vs susceptible by group}
DeltaDelta <- DeltaNotProtected <- DeltaProtected <- fit <- design <- xgroup
for(i in names(xgroup)){
  Subject <- factor(xgroup[[i]]$PATID)
  Outcome <- factor(xgroup[[i]]$Outcome, levels = c(1,0), labels = c("NotProtected","Protected"))
  Timepoint <- factor(xgroup[[i]]$Timepoint, levels = c(0,25), labels = c("baseline","postvax"))
  Age <- xgroup[[i]]$age.vax1
  Sex <- as.factor(xgroup[[i]]$SEX)
  Batch <- factor(xgroup[[i]]$SEQBATCH, levels = c("Aug2019","Nov2019"))
  MALdVax <- factor(xgroup[[i]]$mal.dvax, levels = c(0,1), labels = c("noMALdVAX","yesMALdVAX"))
  CSPAb_baseline <- as.numeric(xgroup[[i]]$pfcsp_pre)
  CSPAb_baseline[is.na(CSPAb_baseline)] <- median(CSPAb_baseline, na.rm = TRUE) #Impute 2 missing values with median of all samples
  #Dfine protective outcome-specific PfSPZ Vaccination effects and append them to the design matrix
  Protected.Postvax <- Outcome == "Protected" & Timepoint=="postvax"
  NotProtected.Postvax <- Outcome =="NotProtected" & Timepoint=="postvax"
  design[[i]] <- model.matrix(~Subject)
  design[[i]] <- cbind(design[[i]], Protected.Postvax, NotProtected.Postvax) #make design matrix according to edgeR 3.5 (p42)
  #https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf
  rownames(design[[i]]) <- xgroup[[i]]$PATID
  print(i)
  print(colnames(design[[i]])) #check colnames
  #remove columns not of full rank
  if(i == "Placebo"){
    # design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0866"]
    # design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0866"]
    # design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0866"]
    }
  if(i == "4.5 x 10^5 PfSPZ"){
    #design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0867"]
    }
  if(i == "9.0 x 10^5 PfSPZ"){
    #design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0771"]
    }
  if(i == "1.8 x 10^6 PfSPZ"){
    #design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0859"]
    #design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0860"]
    #design[[i]] <- design[[i]][, colnames(design[[i]]) != "SubjectK1W0875"]
  }
  ygroup[[i]]     <- estimateDisp(ygroup[[i]],design[[i]], robust = TRUE)
  fit   <- glmQLFit(ygroup[[i]], design[[i]], robust = TRUE)

  #Uninfected/Protected 25 vs 0
  qlf <- glmQLFTest(fit, coef="Protected.Postvax")
  DeltaProtected[[i]] <- topTags(qlf, n = nrow(xgroup[[i]]))$table
  qlf <- glmQLFTest(fit, coef="NotProtected.Postvax")
  DeltaNotProtected[[i]] <- topTags(qlf, n = nrow(xgroup[[i]]))$table
  qlf <- glmQLFTest(fit, contrast = c(rep(0,(ncol(design[[i]])-2)),1,-1))
  DeltaDelta[[i]] <- topTags(qlf, n = nrow(xgroup[[i]]))$table
}
```

### Examine top 15 DEGs in 1.8x10^6 PfSPZ dose group

Δ Protected 

```{r DEG table placebo, echo = FALSE}
knitr::kable(DeltaProtected$`1.8 x 10^6 PfSPZ`[1:15,])
```

Δ Not Protected

```{r DEG table low dose, echo = FALSE}
knitr::kable(DeltaNotProtected$`1.8 x 10^6 PfSPZ`[1:15,])
```

Δ Protected vs. Δ Not Protected

```{r DEG table med dose, echo = FALSE}
knitr::kable(DeltaDelta$`1.8 x 10^6 PfSPZ`[1:15,])
```

Gene-level
1. Identify genes that were differentially induced in uninfected and infected children receiving 1.8x10^6 PfSPZ Vaccine
2. Merge ΔP vs ΔNP, ΔP, and ΔNP tables.
3. Criteria:
    ΔP vs ΔNP P val < 0.005
    ΔP logFC > 0 and ΔP P val < 0.05
    
```{r determine genes of interest, echo=FALSE}
AllDegs1 <- bind_rows(DeltaDelta, .id = "Treatment") %>%
  mutate(Comparison = "DeltaDelta")
AllDegs2 <- bind_rows(DeltaProtected, .id = "Treatment") %>%
  mutate(Comparison = "DeltaProtected")
AllDegs3 <- bind_rows(DeltaNotProtected, .id = "Treatment") %>%
  mutate(Comparison = "DeltaNotProtected")
AllDegs <- bind_rows(AllDegs1, AllDegs2, AllDegs3) %>%
  dplyr::select(Treatment, Comparison, EnsemblID, GeneSymbol, descripton_new, logFC, logCPM, "F", PValue, FDR) %>%
  dplyr::rename(Description = "descripton_new")

gois <- AllDegs %>%
  filter(Treatment == "1.8 x 10^6 PfSPZ") %>%
  droplevels() %>%
  dplyr::select(Treatment, Comparison, EnsemblID, GeneSymbol, Description, logFC, PValue, FDR) %>%
  pivot_wider(., names_from = Comparison, names_glue = "{Comparison}_{.value}", values_from = c(logFC, PValue, FDR)) %>%
  filter(DeltaDelta_PValue < 0.005 & (DeltaDelta_logFC > 0 & DeltaProtected_PValue < 0.05)) %>%
  arrange(DeltaDelta_PValue)

gois_ms <- gois %>%
  dplyr::select(GeneSymbol, DeltaDelta_logFC, DeltaProtected_PValue, DeltaProtected_logFC, DeltaProtected_PValue, DeltaNotProtected_logFC, DeltaNotProtected_PValue)

knitr::kable(gois_ms)
```

### Apply GSEA

Rank genes by -log10(PValue)*sign(logFC). Run fgsea from fgsea package using NamedGeneRankList2GseaTable helper function.

```{r apply fgsea, warning=FALSE, message=FALSE}
set.seed(23)
#restructure dataframes
degtabs <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(ygroup)){
    degtabs[[i]] <- setNames(as.list(c(1:3)), c("DeltaDelta", "DeltaP", "DeltaNP"))
    degtabs[[i]]$DeltaDelta <- DeltaDelta[[i]]
    degtabs[[i]]$DeltaP <- DeltaProtected[[i]]
    degtabs[[i]]$DeltaNP <- DeltaNotProtected[[i]]
}
ranks <- degtabs
for(i in names(degtabs)){
  for(j in names(degtabs[[i]])){
    ranks[[i]][[j]] <-  degtabs[[i]][[j]] %>%
      mutate(rankmetric = -log10(.$PValue)*sign(.$logFC)) %>%
      dplyr::select(GeneSymbol,rankmetric) %>%
      na.omit() %>% 
      distinct() %>% 
      group_by(GeneSymbol) %>%
      summarize(rankmetric = mean(rankmetric)) %>%
      arrange(desc(rankmetric))
  }
}
#Run fgsea from fgsea package using NamedGeneRankList2GseaTable helper function
devtools::source_url("https://github.com/TranLab/ModuleLists/blob/main/NamedGeneRankList2GseaTable.R?raw=TRUE")
GSEAtab <- ranks
for(i in names(ranks)){
  for(j in names(ranks[[i]])){
    GSEAtab[[i]][[j]] <- NamedGeneRankList2GseaTable(rankedgenes = deframe(ranks[[i]][[j]]),
                                                     geneset = "all",
                                                     output_directory = tempdir(),
                                                     filename_prefix = "GSEA",
                                                     sampleSize = 101,
                                                     minSize = 20,
                                                     maxSize = Inf,
                                                     scoreType = "std") %>%
      as_tibble() %>%
      arrange(desc(NES)) %>% 
      dplyr::select(module_type, pathway, ES, NES, size, leadingEdge, pval, padj) %>% 
      mutate(leadingEdge = gsub("^c\\(|\\)$", "", leadingEdge)) %>%
      mutate(leadingEdge = gsub('"', "", leadingEdge)) %>%
      arrange(padj)
  }
}
#bug---last i and j does not run and gives rankmetric as dataframe instead of GSEA output. Temporary solution is to run the last i and j after the loop, which fixes the problem.
GSEAtab$`1.8 x 10^6 PfSPZ`$DeltaNP <- NamedGeneRankList2GseaTable(rankedgenes = deframe(ranks$`1.8 x 10^6 PfSPZ`$DeltaNP),
                                                                  geneset = "all",
                                                                  output_directory = tempdir(),
                                                                  filename_prefix = "GSEA",
                                                                  sampleSize = 101,
                                                                  minSize = 20,
                                                                  maxSize = Inf,
                                                                  scoreType = "std") %>%
  as_tibble() %>%
  arrange(desc(NES)) %>% 
  dplyr::select(module_type, pathway, ES, NES, size, leadingEdge, pval, padj) %>% 
  mutate(leadingEdge = gsub("^c\\(|\\)$", "", leadingEdge)) %>%
  mutate(leadingEdge = gsub('"', "", leadingEdge)) %>%
  arrange(padj)
```

### Visualize GSEA data as bubble plot

Filter based on padj < 0.20.

```{r plot bubble plot, echo = FALSE}
plotDat <- GSEAtab$`1.8 x 10^6 PfSPZ` %>%
  bind_rows(., .id = "comparison") %>%
  mutate(comparison = factor(comparison, levels = c("DeltaP", "DeltaNP", "DeltaDelta"), labels = c("DeltaP", "DeltaNP", "DeltaDelta"))) %>%
  filter(padj < 0.20) %>%
  filter(module_type %in% c("highBTMs", "lowBTMs", "MonacoModules", "BloodGen3Module", "MSigDB_Hallmark_v7.4"))  %>%
  filter(!grepl("TBA", pathway)) %>%
  dplyr::select(comparison, module_type, pathway, leadingEdge, size, NES, padj) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = gsub("Vd2", "Vδ2", pathway)) %>%
  mutate(pathway = gsub("Vg", "Vγ", pathway)) %>%
  mutate(pathway = gsub("HALLMARK_", "", pathway)) %>%
  mutate(pathway = gsub("_", " ", pathway)) %>%
  mutate(pathway = sub(".*?\\_", "", pathway)) %>%
  group_by(module_type) %>%
  mutate(pathway = fct_reorder(pathway, NES, .desc = TRUE)) %>%
  ungroup() %>%
  filter(!grepl("TBD", pathway)) %>%
  mutate(module_type = factor(module_type, levels = c("highBTMs", "MonacoModules", "lowBTMs", "MSigDB_Hallmark_v7.4", "BloodGen3Module"))) %>%
  arrange(desc(neglogpadj)) %>% 
  droplevels()

foodat <- plotDat %>%
  filter(comparison == "DeltaNP")
#plotting options
basetextsize <- 8  
myfont <- "sans"
bubble_max_size <- 6

MyPlot <- plotDat %>%
  filter(module_type %in% c("highBTMs", "MonacoModules"))  %>%
  ggplot(., aes(x = comparison, y = pathway)) +
  geom_point(aes(size=neglogpadj, fill = NES), alpha = 0.65, shape=21, stroke = 0.25) +
  scale_size_area(name = expression(-log[10]~adj.~p~value), max_size = bubble_max_size) +
      scale_fill_gradient2(low = "blue",
                           mid = "white",
                           high = "red") +
  hrbrthemes::theme_ipsum_es(base_family = myfont, base_size = basetextsize) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
facet_wrap(~module_type, scales = "free_y")
```

### Plot GSEA bubbleplots (Figures 4G of pre-print)

Red: enriched in protected/uninfected through 3 months post-vax surveillance
Blue: enriched in not protected/infected through 3 months post-vax surveillance

```{r plot gsea bubbleplot, fig.align='center', fig.width=5, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
addSmallLegend <- function(myPlot, pointSize = 4, textSize = 6, spaceLegend = 0.6) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize))) +
        theme(legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"))
}
print(addSmallLegend(MyPlot))
```