---
title: "KSPZV1 Phagocytosis Analysis"
author: "Tuan M. Tran"
date: "9/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

Perform comparative stats on Jyoti's data.

### Load required packages

```{r load libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
myPath <- "/Volumes/GoogleDrive/My Drive/Tran Lab Shared/Projects/Doris Duke PfSPZ Kenya/Tuan PfSPZ/KenyaPfSPZ/Phagocytosis Experiments/"
```

### Load data

```{r load data, echo=FALSE, message=FALSE, warning=FALSE}
countdat <- readRDS(paste0(myPath,"PfSPZ_phagocytosis_allExperiments_Countdata_new_09212021.rds"))
freqdat <- readRDS(paste0(myPath,"PfSPZ_phagocytosis_allExperiments_Freqdata_new1_09212021.rds"))
```

```{r clean data and correct inconsistent naming, echo=FALSE, include=FALSE}
freqdat_clean <- freqdat %>%
  mutate(Treatment = gsub("Unstimulated", "vehicle", Treatment)) %>%
  mutate(Treatment = gsub("B-glucan", "β-glucan", Treatment)) %>%
  mutate(Treatment = gsub("Hemozoin", "sHz", Treatment)) %>%
  mutate(Treatment = gsub("Flagellin", "flagellin", Treatment)) %>%
  mutate(Treatment = gsub("PolyIC", "poly I:C", Treatment)) %>%
  mutate(Treatment = gsub("PfIRBC", "PfiRBC", Treatment)) %>%
  mutate(Treatment = gsub("No spz control", "no spz control", Treatment)) %>%
  mutate(Treatment = gsub("Imiquimod", "imiquimod", Treatment)) %>%
  mutate(Treatment = factor(Treatment, levels = c("vehicle", "β-glucan", "CpG","LPS", "poly I:C", "flagellin", "imiquimod","sHz", "uRBC", "PfiRBC", "no spz control")))
```

```{r plot freq, echo = FALSE, include=FALSE}
mycomparisons <- list( c("vehicle", "β-glucan"), c("vehicle", "CpG"), c("vehicle", "LPS"), c("vehicle", "poly I:C"), c("vehicle", "flagellin"), c("vehicle", "imiquimod"),
                       c("vehicle", "sHz"), c("vehicle", "uRBC"), c("vehicle", "PfiRBC"))

freqdat_clean %>%
  group_by(group, Treatment, Experiment, Gate) %>%
  mutate(Frequency = mean(Frequency)) %>%
  group_by(Experiment, Treatment, Gate) %>%
  mutate(Frequency = mean(Frequency)) %>%
  ungroup() %>%
  ggplot(., aes(x = Treatment, y = Frequency)) +
  geom_boxplot() +
  geom_jitter() +
  stat_compare_means(comparisons = mycomparisons) + 
  stat_compare_means() +
  hrbrthemes::theme_ipsum() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_wrap(~Experiment+Gate, scales = "free_y", ncol = 3) 
```

```{r recalc freq, echo = FALSE, include=FALSE}
foodat <- countdat %>%
  mutate(Treatment = gsub("Unstimulated", "vehicle", Treatment)) %>%
  mutate(Treatment = gsub("B-glucan", "β-glucan", Treatment)) %>%
  mutate(Treatment = gsub("Hemozoin", "sHz", Treatment)) %>%
  mutate(Treatment = gsub("Flagellin", "flagellin", Treatment)) %>%
  mutate(Treatment = gsub("Poly IC", "poly I:C", Treatment)) %>%
  mutate(Treatment = gsub("PfIRBC", "PfiRBC", Treatment)) %>%
  mutate(Treatment = gsub("No spz control", "no spz control", Treatment)) %>%
  mutate(Treatment = gsub("Imiquimod", "imiquimod", Treatment)) %>%
  mutate(Treatment = factor(Treatment, levels = c("vehicle", "β-glucan", "CpG","LPS", "poly I:C", "flagellin", "imiquimod","sHz", "uRBC", "PfiRBC", "no spz control"))) %>%
  mutate(Replicate = gsub(" ", "", Replicate)) %>%
  # mutate(Gate = gsub("\\+", "_pos", Gate)) %>%
  # mutate(Gate = gsub(" ", "_", Gate)) %>%
  # mutate(Gate = gsub("\\-", "_neg", Gate)) %>%
  # mutate(Gate = gsub("\\(", "_", Gate)) %>%
  # mutate(Gate = gsub("\\)", "", Gate)) %>%
  # mutate(Gate = make.names(Gate)) %>%
  pivot_wider(., names_from = Gate, values_from = Count) %>%
  dplyr::select(-c(`% Phagocytosis`)) %>%
  mutate(across(c(5:ncol(.)), .fns = ~1000*(./`Hoechst+`))) %>%
  pivot_longer(., cols = 5:ncol(.), names_to = "Gate", values_to = "Frequency")
```

```{r plot freq 2, echo = FALSE, include=FALSE}
mycomparisons <- list( c("vehicle", "β-glucan"), c("vehicle", "CpG"), c("vehicle", "LPS"), c("vehicle", "poly I:C"), c("vehicle", "flagellin"), c("vehicle", "imiquimod"),
                       c("vehicle", "sHz"), c("vehicle", "uRBC"), c("vehicle", "PfiRBC"))

plotdat <- foodat %>%
  filter(Gate %in% c("Hoechst+Py+Total", "Hoechst+Py+Internal", "Hoechst+Py+External")) %>%
  dplyr::select(-c(group)) %>%
  group_by(Replicate, Treatment, Experiment, Gate) %>%
  summarise(Frequency = mean(Frequency)) %>%
  ungroup()
plotdat %>%
  ggplot(., aes(x = Treatment, y = Frequency)) +
  geom_boxplot() +
  geom_jitter() +
  stat_compare_means(label = "p.signif", method = "t.test",
                     ref.group = "vehicle", paired = FALSE) + 
  stat_compare_means(method = "anova", label.y = 1) +
  ylab("Counts per 1000 monocytes") +
  hrbrthemes::theme_ipsum() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_wrap(~Gate, scales = "free_y", ncol = 5) 

plotdat %>%
  filter(Treatment %in% c("vehicle", "poly I:C")) %>%
  filter(Experiment != 4) %>%
  ggpaired(., x = "Treatment", y = "Frequency",
         color = "Treatment", line.color = "gray", line.size = 0.4,
         palette = "jco")+
  stat_compare_means(method = "t.test", paired = TRUE) +
  facet_wrap(~Gate, scales = "free_y", ncol = 5) 
```

### Use proportion approach

1. determine which experiments were assigned to each treatment
2. sum up all replicates for all experiments in which a treatment was used for both treatment and vehicle
  a. include only vehicle replicates within experiments of treatment of interest
3. annotate treatments with experiment information
4. use Fisher's exact to find differences in proportions between Hoechst+Py+Internal and Hoechst+ for treatment of interest and vehicle
5. extract OR, 95% CIs, and p values from the FET analysis
6. determine BH-adjusted P values

```{r arrange for proportional testing by experiments, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
propdat <- countdat %>%
  mutate(Treatment = gsub("Unstimulated", "vehicle", Treatment)) %>%
  mutate(Treatment = gsub("B-glucan", "beta-glucan", Treatment)) %>%
  mutate(Treatment = gsub("B-Glucan", "beta-glucan", Treatment)) %>%
  #mutate(Treatment = gsub("Hemozoin", "sHz", Treatment)) %>%
  mutate(Treatment = gsub("Flagellin", "flagellin", Treatment)) %>%
  mutate(Treatment = gsub("Poly IC", "poly I:C", Treatment)) %>%
  mutate(Treatment = gsub("PfIRBC", "PfiRBC", Treatment)) %>%
  mutate(Treatment = gsub("No spz control", "no spz control", Treatment)) %>%
  mutate(Treatment = gsub("Imiquimod", "imiquimod", Treatment)) %>%
  mutate(Treatment = factor(Treatment, levels = c("vehicle", "beta-glucan", "CpG","LPS", "poly I:C", "flagellin", "imiquimod","sHz", "uRBC", "PfiRBC", "no spz control"))) %>%
  mutate(Replicate = gsub(" ", "", Replicate)) %>%
  mutate(Experiment = factor(Experiment)) %>%
  dplyr::select(-c(`% Phagocytosis`)) %>%
  filter(Gate %in% c("Hoechst+Py+Total", "Hoechst+Py+Internal", "Hoechst+Py+External", "Hoechst+"))  %>%
  filter(Treatment %in% c("vehicle", "beta-glucan", "CpG","LPS", "poly I:C", "flagellin", "imiquimod","sHz", "uRBC", "PfiRBC"))  %>%
  dplyr::select(-c(group, Replicate)) %>%
  group_by(Experiment, Treatment, Gate) %>%
  summarise(Count = sum(Count)) %>%
  ungroup() %>%
  droplevels()

fetdat <- tempdat <- mydat <- mydat_bound <- fetres <- Exp_annot <- c()
for(i in c("beta-glucan", "CpG","LPS", "poly I:C", "flagellin", "imiquimod","sHz", "uRBC", "PfiRBC")){
  tempdat[[i]] <- propdat[propdat$Treatment == i,] %>%
    droplevels
   for(j in unique(tempdat[[i]]$Experiment)){
    mydat[[i]][[j]] <- propdat[propdat$Treatment %in% c("vehicle", i) & propdat$Experiment==j,] %>%
      dplyr::select(-c(Experiment)) %>%
      mutate(Experiments = paste(unique(tempdat[[i]]$Experiment), collapse=",")) %>%
      mutate(n_Experiments = length(unique(tempdat[[i]]$Experiment)))
   }
  mydat_bound[[i]] <- bind_rows(mydat[[i]], .id="Experiment") %>%
    dplyr::select(-c(Experiment)) %>%
    group_by(Treatment, Gate, Experiments, n_Experiments) %>%
    summarise(Count = sum(Count)) %>%
    ungroup() %>%
    droplevels()
  Exp_annot[[i]]  <- bind_rows(mydat[[i]], .id="Experiment") %>%
    dplyr::select(Treatment, Experiments, n_Experiments) %>%
    filter(Treatment == i) %>%
    distinct(Treatment, .keep_all = TRUE)
  fetres[[i]] <- fisher.test(xtabs(Count ~ Treatment+Gate, data=mydat_bound[[i]])[,c("Hoechst+","Hoechst+Py+Internal")])
  fetdat[[i]] <- data.frame("OR" = fetres[[i]]$estimate, "LCI" = fetres[[i]]$conf.int[1], "UCI" = fetres[[i]]$conf.int[2], "p" = fetres[[i]]$p.value)
}

all_annots <- bind_rows(Exp_annot, .id = "Treatment")
allfets <- bind_rows(fetdat, .id = "Treatment") %>%  
  as_tibble() %>%
  filter(!Treatment %in% c("PfiRBC", "sHz", "uRBC")) %>%
  droplevels() %>%
  mutate(BH.adj.p = p.adjust(.$p, n = nrow(.))) %>%
  mutate(significance = ifelse(BH.adj.p<0.0001, "****",
                              ifelse(BH.adj.p<0.001, "***",
                                     ifelse(BH.adj.p<0.01, "**",
                                            ifelse(BH.adj.p<0.05, "*", "ns"))))) %>%
  left_join(., all_annots, by = "Treatment") %>%
  mutate(Treatment = gsub("beta", "β", Treatment)) %>%
  mutate(label = paste0(Treatment, " (", n_Experiments, ")")) %>%
  mutate("BH-adjusted p value" = BH.adj.p) %>%
  mutate("Treatment (n experiments)" = label)
```

### Experiment

THP-1 cells were treated with indicated agonist or media vehicle for 22-36h and then live P. yoelii sporozoites (dissected from salivary glands) were added and co-incubated for 1 h at 37C. Phagocytosis was mesaured via ImageStream flow cytometry using intracellular staining with the 2F6 anti-PyCSP mAb and Hoescht.

### Results in tabular form

```{r plot results in table, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(allfets %>%
               mutate(across(c(2:6,11), scales::scientific, 3)) %>%
               dplyr::select(12,2:5,11))
knitr::kable(allfets %>%
               mutate(across(c(2:6,11), scales::scientific, 3)) 

```

### Results as a forest plot

1. Reorder treatment/labels by descending BH-adjusted P value
2. add # of experiments in parenthesis

```{r plot forest plot, echo = FALSE, fig.align='center',fig.width=6, fig.height=4}
myForestPlot <- allfets %>%
  mutate(label = fct_reorder(label, BH.adj.p, .desc = TRUE)) %>%
  mutate(Treatment = fct_reorder(label, BH.adj.p, .desc = TRUE)) %>%
  ggplot(aes(y=label, x=OR)) +
  geom_point(size=2, shape=19) +
  geom_errorbarh(aes(xmin=LCI, xmax=UCI), height=.3) +
  coord_fixed(ratio=.3) +
  geom_vline(xintercept=1, linetype='longdash') +
  ylab("Treatment (n experiments") +
  xlab("Odds Ratio") +
  theme_bw()
print(myForestPlot)
```


```{r, print Forest Plot, echo=FALSE}
cairo_pdf(filename = "/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Figures/Working Figures for Submission/Fig 6D THP1 Phagocytosis Forest Plot.pdf", height = 3.5, width = 2.5)
print(myForestPlot)
dev.off()
```





