---
title: "KSPZV1 DGE pre-immunization baseline - revised"
author: "Tuan M. Tran"
date: "01/31/2021"
output: html_document
---

```{r libraries, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(edgeR)
library(googledrive)
library(tidyverse)
library(fgsea)
library(data.table)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective

Perform differential gene expression using edgeR on pre-immunization baseline whole-blood samples from the KSPZV1 malaria vaccine trial. 
Revised analysis includes adjustments for:
1. Study site
2. CSP Ab
3. parasitemia at VAX1 (closest to baseline)

### Load ExpressionSet

```{r load SeqExpressionSet 2, message=FALSE, warning=FALSE}
#from google drive
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("1Q4VXYxdl9CqtzhcvHA9pn8q06_YOZgi8"), path = temp, overwrite = TRUE)
x <- readRDS(file = dl$local_path)
x$treat <- factor(x$treat, levels = c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
```

### Set options

```{r set options, message=FALSE, warning=FALSE}
myTimepoint <- 0
```

### Reduce samples

```{r reduced data, message=FALSE, warning=FALSE}
xgroup <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(xgroup)){
  xgroup[[i]] <- x[, which(x$treat == i)]
  xgroup[[i]] <- xgroup[[i]][,xgroup[[i]]$Timepoint == myTimepoint]
  xgroup[[i]]$treat <- droplevels(xgroup[[i]]$treat)
  #impute only missing CSP Ab value as median CSP A of all baseline samples (only affects one subject in 9.0x10^6 PfSPZ dose group)
  xgroup[[i]]$CSPAb <- ifelse(is.na(xgroup[[i]]$CSPAb), median(xgroup[[i]]$CSPAb, na.rm = TRUE), xgroup[[i]]$CSPAb)
  print(i)
  print(table(Outcome = xgroup[[i]]$Outcome, Timepoint = xgroup[[i]]$Timepoint))
  print(table(Outcome = xgroup[[i]]$Outcome, Dosegroup = xgroup[[i]]$treat))
  print(dim(xgroup[[i]]))
}
```

### Build DGEList Object

```{r make DGEList object, message=FALSE, warning=FALSE}
#Define group by dose, timepoint and outcome
ygroup <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(ygroup)){
  ygroup[[i]]  <- DGEList(counts=counts(xgroup[[i]]), genes=fData(xgroup[[i]]), group= factor(paste(xgroup[[i]]$treat, xgroup[[i]]$Timepoint, xgroup[[i]]$Outcome, sep = "_")), remove.zeros=T)
  ygroup[[i]]$samples$PATID <- gsub("_.*", "", rownames(ygroup[[i]]$samples))
}
```

### Filter out low expression features

```{r low expression filters, message=FALSE, warning=FALSE}
for(i in names(ygroup)){
  keep <- filterByExpr(ygroup[[i]])
  ygroup[[i]] <- ygroup[[i]][keep, , keep.lib.sizes=FALSE]
  ygroup[[i]]$samples$lib.size <- colSums(ygroup[[i]]$counts)
  ygroup[[i]] <- calcNormFactors(ygroup[[i]])	#Normalization
}
```

### Analysis between Not Protected and Protected within a Dose Group


```{r, build design matrix protected vs not protected by group, message=FALSE, warning=FALSE}
degtab <- setNames(as.list(c(1:4)), c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))
for(i in names(ygroup)){
  Subject <- factor(xgroup[[i]]$PATID)
  Outcome <- factor(xgroup[[i]]$Outcome, levels = c(1,0), labels = c("NotProtected","Protected"))
  Timepoint <- factor(xgroup[[i]]$Timepoint, levels = c(0,25), labels = c("baseline","postvax"))
  Batch <- factor(xgroup[[i]]$SEQBATCH, levels = c("Aug2019","Nov2019"))
  Site <- as.factor(xgroup[[i]]$site)
  CSPAb <- as.integer(xgroup[[i]]$CSPAb)
  Pf_VAX1 <- as.factor(xgroup[[i]]$mal.vax.1)
  #This is the revised Timepoint 0 analysis.
  design <- model.matrix(~Batch + CSPAb + Site + Pf_VAX1 + Outcome)
  print(i)
  print(colnames(design))
  rownames(design) <- ygroup[[i]]$samples$PATID
  ygroup[[i]]     <- estimateDisp(ygroup[[i]],design, robust = TRUE)
  fit   <- glmQLFit(ygroup[[i]], design, robust = TRUE)
  qlf <- glmQLFTest(fit, contrast = c(0,0,0,0,0,1))
  degtab[[i]] <- topTags(qlf, n = nrow(ygroup[[i]]))$table
}

temp <- degtab
for(i in names(degtab)){
  temp[[i]] <- degtab[[i]] %>%
    dplyr::select(7,8,10:14) %>%
    head(.,15)
  }
```

### Examine top 15 DEGs (protected vs not protected)

Placebo 

```{r DEG table placebo, echo = FALSE}
knitr::kable(temp$Placebo)
```

4.5 x 10^5 PfSPZ 

```{r DEG table low dose, echo = FALSE}
knitr::kable(temp$`4.5 x 10^5 PfSPZ`)
```

9.0 x 10^5 PfSPZ 

```{r DEG table med dose, echo = FALSE}
knitr::kable(temp$`9.0 x 10^5 PfSPZ`)
```

1.8 x 10^6 PfSPZ 

```{r DEG table high dose, echo = FALSE}
knitr::kable(temp$`1.8 x 10^6 PfSPZ`)
```

### Apply GSEA

Rank genes by -log10(PValue)*sign(logFC). Run fgsea from fgsea package using NamedGeneRankList2GseaTable helper function.

```{r apply fgsea, warning=FALSE, message=FALSE}
set.seed(23)
ranks <- degtab
for(i in names(degtab)){
  ranks[[i]] <- degtab[[i]] %>% 
  mutate(rankmetric = -log10(.$PValue)*sign(.$logFC)) %>%
  dplyr::select(GeneSymbol,rankmetric) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(GeneSymbol) %>%  
  summarize(rankmetric = mean(rankmetric)) %>%
  arrange(desc(rankmetric)) %>%
  deframe()
}
#Run fgsea from fgsea package using NamedGeneRankList2GseaTable helper function
devtools::source_url("https://github.com/TranLab/ModuleLists/blob/main/NamedGeneRankList2GseaTable.R?raw=TRUE")
GSEAtab <- ranks
for(i in names(ranks)){
  GSEAtab[[i]] <- NamedGeneRankList2GseaTable(rankedgenes = ranks[[i]],
                                         geneset = "all",
                                         output_directory = tempdir(),
                                         filename_prefix = "GSEA",
                                         sampleSize = 101,
                                         minSize = 20,
                                         maxSize = Inf,
                                         scoreType = "std") %>%
    as_tibble() %>%
    arrange(desc(NES)) %>% 
    dplyr::select(module_type, pathway, ES, NES, size, leadingEdge, pval, padj) %>% 
    mutate(leadingEdge = gsub("^c\\(|\\)$", "", leadingEdge)) %>%
    mutate(leadingEdge = gsub('"', "", leadingEdge)) %>%
    arrange(padj)
}
# These are the genes with duplicated rankmetrics...can ignore warning about ties b/c it's trivial (only involves two genes)
# 1 RP11-521C22.3    -0.0303
# 2 AC069547.1       -0.0303
```


### Visualize GSEA data as bubble plot

Filter based on padj < 0.20.

```{r visualize GSEA data, message=FALSE, warning=FALSE}
plotDat <- GSEAtab %>%
  bind_rows(., .id = "dosegroup") %>%
  mutate(dosegroup = factor(dosegroup, levels = c("Placebo", "4.5 x 10^5 PfSPZ", "9.0 x 10^5 PfSPZ", "1.8 x 10^6 PfSPZ"))) %>%
  filter(padj < 0.20) %>%
  filter(module_type %in% c("highBTMs", "lowBTMs", "MonacoModules", "BloodGen3Module", "MSigDB_Hallmark_v7.4"))  %>%
  filter(!grepl("TBA", pathway)) %>%
  dplyr::select(dosegroup, module_type, pathway, leadingEdge, size, NES, padj) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = gsub("Vd2", "Vδ2", pathway)) %>%
  mutate(pathway = gsub("Vg", "Vγ", pathway)) %>%
  mutate(pathway = gsub("HALLMARK_", "", pathway)) %>%
  mutate(pathway = gsub("_", " ", pathway)) %>%
  mutate(pathway = sub(".*?\\_", "", pathway)) %>%
  group_by(module_type, dosegroup) %>%
  mutate(pathway = fct_reorder(pathway, NES, .desc = TRUE)) %>%
  ungroup() %>%
  mutate(dosegroup = fct_rev(dosegroup)) %>%
  filter(!grepl("TBD", pathway)) %>%
  mutate(module_type = factor(module_type, levels = c("highBTMs", "MonacoModules", "lowBTMs", "MSigDB_Hallmark_v7.4", "BloodGen3Module"))) %>%
  arrange(desc(neglogpadj)) %>% 
  droplevels()

#plotting options
basetextsize <- 8  
myfont <- "sans"
bubble_max_size <- 6

TopPlot <- plotDat %>%
  filter(module_type %in% c("highBTMs", "MonacoModules"))  %>%
  ggplot(., aes(x = pathway, y = dosegroup)) +
  geom_point(aes(size=neglogpadj, fill = NES), alpha = 0.65, shape=21, stroke = 0.25) +
  scale_size_area(name = expression(-log[10]~adj.~p~value), max_size = bubble_max_size) +
      scale_fill_gradient2(low = "blue",
                           mid = "white",
                           high = "red") +
  hrbrthemes::theme_ipsum_es(base_family = myfont, base_size = basetextsize) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
facet_wrap(~module_type, scales = "free_x", nrow= 2)

MidPlot <- plotDat %>%
  filter(module_type %in% c("lowBTMs", "MSigDB_Hallmark_v7.4"))  %>%
  ggplot(., aes(x = pathway, y = dosegroup)) +
  geom_point(aes(size=neglogpadj, fill = NES), alpha = 0.65, shape=21, stroke = 0.25) +
  scale_size_area(name = expression(-log[10]~adj.~p~value), max_size = bubble_max_size) +
      scale_fill_gradient2(low = "blue",
                           mid = "white",
                           high = "red") +
  hrbrthemes::theme_ipsum_es(base_family = myfont, base_size = basetextsize) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
facet_wrap(~module_type, scales = "free_x")

BottomPlot <- plotDat %>%
  filter(module_type %in% c("BloodGen3Module"))  %>%
  ggplot(., aes(x = pathway, y = dosegroup)) +
  geom_point(aes(size=neglogpadj, fill = NES), alpha = 0.65, shape=21, stroke = 0.25) +
  scale_size_area(name = expression(-log[10]~adj.~p~value), max_size = bubble_max_size) +
      scale_fill_gradient2(low = "blue",
                           mid = "white",
                           high = "red") +
  hrbrthemes::theme_ipsum_es(base_family = myfont, base_size = basetextsize) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
facet_wrap(~module_type, scales = "free_x")
```

### Plot GSEA bubbleplots (Figures 2D and S3 of revised manuscript)

Red: enriched in protected/uninfected through 3 months post-vax surveillance
Blue: enriched in not protected/infected through 3 months post-vax surveillance

```{r plot gsea bubbleplot, fig.align='center', fig.width=6.125, fig.height=5.55, echo=FALSE, message=FALSE, warning=FALSE}
print(TopPlot)
```

```{r plot gsea bubbleplot 2, fig.align='center', fig.width=12.5, fig.height=4.5, echo=FALSE, message=FALSE, warning=FALSE}
print(MidPlot)
```

```{r plot gsea bubbleplot 3, fig.align='center', fig.width=12, fig.height=3.75, echo=FALSE, message=FALSE, warning=FALSE}
print(BottomPlot)
```


```{r save DEG tables, echo=FALSE, message=FALSE, warning=FALSE}
# degtab_long <- bind_rows(degtab, .id = "treatment") %>%
#   dplyr::select(treatment, GeneSymbol, EnsemblID, descripton_new, logFC,logCPM, F, PValue, FDR) %>%
#   mutate(study = "KSPZV1", timepoint = "pre-immunization baseline adj batch, CSP Ab baseline, Pf VAX1, site", comparison = "protected (P) vs. not protected (NP) at 3 months") %>%
#   dplyr::select(study, treatment, timepoint, comparison, GeneSymbol, EnsemblID, descripton_new, logFC,logCPM, F, PValue, FDR) %>%
#   dplyr::rename(hgnc_symbol = GeneSymbol, ensembl_id = EnsemblID, description = descripton_new)
# degtab_VRC <- readxl::read_excel("/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Cell Submission October 2021/Supplementary Tables/Table S1 Prevax DEG Tables for KSPZV1 VRC312 VR314.xlsx") %>%
#               dplyr::filter(study == "VRC312" | study == "VRC314")
#   bind_rows(degtab_long, degtab_VRC) %>%
#     as_tibble() %>%
#     writexl::write_xlsx("/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Supplementary Tables/Table S1 Prevax DEG Tables for KSPZV1 VRC312 VR314 revision.xlsx")
```

```{r save GSEA tables, echo=FALSE, message=FALSE, warning=FALSE}
# bind_rows(GSEAtab, .id = "treatment") %>%
#   as_tibble() %>%
#   writexl::write_xlsx("/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Supplementary Tables/Table SX Prevax GSEA Tables for KSPZV1 revision.xlsx")
```

```{r save plots 1, echo=FALSE, message=FALSE, warning=FALSE}
#Figure 2D high-annotation level BTM and Monaco modules
# Cairo::CairoPDF(file = "/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Figures/Fig2D revised hiBTM Monaco GSEA Bubble Plot.pdf",
#                 width = 6.125, height = 5.65)
# TopPlot
# dev.off()
```

```{r save plots 2, echo=FALSE, message=FALSE, warning=FALSE}
#Figure S3A,B low-annotation level BTM. Hallmarkmodules
# FigS3_plot <- ggpubr::ggarrange(MidPlot, BottomPlot, common.legend = TRUE, nrow=2)
# ggsave("/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Figures/Fig S3ABC revised lowBTM HALLMARK BloodGen3 GSEA Bubble Plot.pdf", FigS3_plot, device = "pdf", units = "in", height = 8, width = 14)
```

```{r save for IPA, echo = FALSE}
tabdir <- "/Volumes/GoogleDrive/My Drive/Manuscripts/KSPZV1 Manuscript/Med Revised Submission 2022/Supplementary Tables/"

temp <- degtab
for(i in names(degtab)){
  temp[[i]] <- degtab[[i]] %>%
    dplyr::select(Geneid, EnsemblID, logFC, PValue, FDR)
  }

temp$Placebo %>%
   write_csv(paste0(tabdir, "Table S1 Prevax DEG KSPZV1 rev placebo IPA.csv"))
temp$`4.5 x 10^5 PfSPZ` %>%
   write_csv(paste0(tabdir, "Table S1 Prevax DEG KSPZV1 rev lowdose IPA.csv"))
temp$`9.0 x 10^5 PfSPZ` %>%
   write_csv(paste0(tabdir, "Table S1 Prevax DEG KSPZV1 rev meddose IPA.csv"))
temp$`1.8 x 10^6 PfSPZ` %>%
   write_csv(paste0(tabdir, "Table S1 Prevax DEG KSPZV1 rev highdose IPA.csv"))
```

