---
title: "KSPZV1 WGCNA Delta Postvax"
author: "Tuan M. Tran"
date: "12/1/2022"
output: html_document
---

### Objective

Perform weighted gene correlation network analysis as originally described by Horvath et al. on delta (postvax/baseline) transcriptomic data from the KSPZV1 clinical trial.

References:

https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/
https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/JMiller/Tutorial%20document.pdf

### Load packages

```{r load packages, message=FALSE, warning=FALSE}
library(edgeR)
library(readxl)
library(EDASeq)
library(Biobase)
library(WGCNA)
library(ape)
library(Cairo)
library(CorLevelPlot)
library(tidyverse)
library(igraph)
library(remotes)
library(fgsea)
library(data.table)
library(ggplot2)
library(viridis)
library(ggpubr)
library(googledrive)
allowWGCNAThreads()
```

### Options and define variables

```{r options and define variables, warning=FALSE, message=FALSE}
myCor <- "cor"
power <- 11.5
myMergeCutHeight <- 0.05 
myDeepSplit <- 2 
minModSize <- 20 
enforceMMS <- FALSE
cor.pval <- 0.05
```

### Load ExpressionSet

```{r load cpm eset, warning=FALSE, message=FALSE}
#local file: "PfSPZ_cpm_ExpressionSet_230x23768_allGroups_bothBatches_delta_rmBatchFX_06082021_TMT.rds"

temp <- tempfile(fileext = ".rds")
dl <- drive_download(
  as_id("17xx0YaggLiyWdJc9rqA1nkPSE7dj05p0"), path = temp, overwrite = TRUE)
x <- readRDS(file = dl$local_path)
dim(x)
```

### Make weighted gene correlation matrix based on full data set

```{r weighted gene correlation matrix based on full data set, message=FALSE, warning=FALSE}
WGCNA_matrix <- t(exprs(x))
blockSize(ncol(WGCNA_matrix), rectangularBlocks = TRUE, maxMemoryAllocation = 4^31)
par(mfrow=c(1,1))
plotClusterTreeSamples(datExpr=WGCNA_matrix)
powers <- seq(4,20,by=0.5)
sft <- pickSoftThreshold(WGCNA_matrix, powerVector = powers, corFnc = myCor, verbose = 5, networkType ="signed", blockSize = ncol(WGCNA_matrix))
sft$powerEstimate
par(mfrow=c(1,1))
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab='Soft Threshold (power)',ylab='Scale Free Topology Model Fit,signed R²',
     type='n', main = paste('Scale independence'));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=1,col='red'); abline(h=0.90,col='red')
```

### Run an automated network analysis

```{r calculation of adjacency matrix, message=FALSE, warning=FALSE}
net <- blockwiseModules(WGCNA_matrix,
                        power=power,
                        deepSplit=myDeepSplit,
                        minModuleSize=minModSize,
                        TOMType="none", 
                        mergeCutHeight=myMergeCutHeight,
                        TOMDenom="mean",
                        detectCutHeight=0.995,
                        corType="pearson",
                        networkType="signed",
                        pamStage=TRUE,
                        pamRespectsDendro=TRUE,
                        reassignThresh=0.05,
                        verbose=5,
                        saveTOMs=FALSE,
                        maxBlockSize=ncol(WGCNA_matrix),
                        nThreads = 0)

# Summary Module Table
nModules <- length(table(net$colors))-1
modules <- cbind(colnames(as.matrix(table(net$colors))),table(net$colors))
orderedModules <- cbind(Mnum=paste("M",seq(1:nModules),sep=""),Color=labels2colors(c(1:nModules)))
modules <- modules[match(as.character(orderedModules[,2]),rownames(modules)),]
tmpMEs <- MEs <- net$MEs
colnames(tmpMEs) <- paste("ME",colnames(MEs),sep="")
kMEdat <- signedKME(WGCNA_matrix, tmpMEs, corFnc=myCor) #calculate (signed) eigengene-based connectivity, also known as module membership
WGCNA_dat <- cbind(fData(x)$GeneSymbol, colnames(WGCNA_matrix),net$colors,kMEdat) %>%
  as.data.frame() %>%
  dplyr::rename(GeneSymbol = "fData(x)$GeneSymbol") %>%
  dplyr::rename(ENSEMBLID = "colnames(WGCNA_matrix)") %>%
  dplyr::rename(ModuleColors = "net$colors")
```

### Correlate modules with traits

```{r correlate modules with traits, message=FALSE, warning=FALSE}
# Define numbers of genes and samples
nGenes = ncol(WGCNA_matrix)
nSamples = nrow(WGCNA_matrix)

datvar <- pData(x) %>%
  dplyr::select(PATID, SEQBATCH, site, SEX, age.vax1, mal.vax.1, treat, mal.atp.3, tte.mal.atp.6, mal.dvax, mal.dvax.tot, pfcsp_pre, pfcsp_post, log2FC_CSPAb) %>%
  mutate('Gender (female)' = factor(ifelse(SEX == "F", 1, 0))) %>%
  mutate('Site (Siaya)' = factor(ifelse(site == "Siaya", 1, 0))) %>%
  dplyr::rename('Pf infection at first vax' = "mal.vax.1") %>%
  mutate('pre-vax anti-CSP IgG' = log2(pfcsp_pre+1)) %>%
  dplyr::rename('log2FC anti-CSP IgG' = "log2FC_CSPAb") %>%
  mutate('1.8 x 10^6 PfSPZ' = factor(ifelse(treat == "1.8 x 10^6 PfSPZ", 1, 0))) %>%
  mutate('parasitemic events during vax period' = mal.dvax.tot) %>%
  mutate('uninfected, 3 months' = factor(ifelse(mal.atp.3 == 0, 1, 0))) %>%
  dplyr::rename('days to first parasitemia' = "tte.mal.atp.6") %>%
  mutate(Age = age.vax1) %>%
  dplyr::select(PATID, Age, 'pre-vax anti-CSP IgG', 'parasitemic events during vax period', 'uninfected, 3 months', 'days to first parasitemia', 'log2FC anti-CSP IgG') %>% #delta
  as_tibble() %>%
  column_to_rownames(var = "PATID") %>%
  mutate_if(is.factor, as.character) %>%
  mutate_if(is.character, as.numeric)

modTraitCor <- cor(orderMEs(net$MEs), datvar, use = "p")
modTraitP <- corPvalueStudent(modTraitCor, nSamples)
#Since we have a moderately large number of modules and traits, a suitable graphical representation will help in reading the table. We color code each association by the correlation value: Will display correlations and their p-values

#Select out only modules that have P<0.05 in Protection
modTraitP.temp <- modTraitP %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module") %>%
  filter(.$'uninfected, 3 months' < cor.pval | .$'days to first parasitemia' < cor.pval)
modTraitCor.select <- modTraitCor[modTraitP.temp$Module,]
modTraitP.select <- modTraitP[modTraitP.temp$Module,]
textMatrix <- paste(signif(modTraitCor.select, 2), "\n(P=",
                   signif(modTraitP.select, 1), ")", sep = "")
dim(textMatrix) <- dim(modTraitCor.select)
```

### Display the correlation values within a heatmap plot (Figure 4E of pre-print)

```{r plot module trait correlations heatmap, warning=FALSE, message=FALSE, fig.align='center', fig.width=8, fig.height=8}
par(mar = c(11, 9, 1, 1))
labeledHeatmap(Matrix = modTraitCor.select, xLabels = names(datvar),
               yLabels = rownames(modTraitCor.select), ySymbols = rownames(modTraitCor.select),
               colorLabels =FALSE,colors=blueWhiteRed(100),textMatrix=textMatrix,
               setStdMargins = FALSE, zlim = c(-1,1),
               main = paste("Module-trait relationships"),xLabelsAngle = 45) 
```

### Identify hub genes

```{r id hub genes, warning=FALSE, message=FALSE}
myColors <- gsub("ME", "", rownames(modTraitCor.select))
topmodules <- chooseTopHubInEachModule(WGCNA_matrix, net$colors, omitColors = "grey", power = 12.5, type ="signed")
mytopmodules <- topmodules[myColors] %>%
  as.data.frame() %>%
  dplyr::rename(EnsemblID = ".") %>%
  rownames_to_column("module_label") %>%
  as_tibble() %>%
  left_join(., fData(x) %>%
              dplyr::select(EnsemblID, GeneSymbol), by = "EnsemblID")
devtools::source_url("https://github.com/jtlovell/limmaDE2/blob/master/R/wgcna2igraph.R?raw=TRUE")
graph <- wgcna2igraph(net = net, WGCNA_matrix, modules2plot = myColors,
                      colors2plot = myColors,
                      kME.threshold = 0.5, adjacency.threshold = 0.1,
                      adj.power = power, verbose = T,
                      node.size = 1, frame.color = NA, node.color = scales::muted("red"),
                      edge.alpha = .7, edge.width = 0.5)
hubscores <- hub_score(graph, scale = TRUE, weights = NULL,
  options = arpack_defaults)
```

### Top modules and hub genes (rownames for Figure 4E of pre-print)

```{r display my topmodules, echo=FALSE}
knitr::kable(mytopmodules)
```

### Plot network graph of significant modules (Figure 4F of pre-print)

Network graphs of significant modules containing nodes (red dots) and edges (lines) meeting minimum thresholds. Correlations between nodes in different modules are shown as black edges.

```{r plot networkd graph, message=FALSE, warning=FALSE, fig.align='center', fig.width=8, fig.height=8, echo=FALSE}
plot(graph)
```

### Identify modules with < 0.05

List number of genes per module

```{r id modules that are significant than pull out gene list, warning=FALSE, message=FALSE}
WGCNA_dat_select <- c()
for(i in modTraitP.temp$Module){
  module.color <- sub("ME","", i)
  module.colname <- paste0("kME", i)
  WGCNA_dat_select[[i]] <- WGCNA_dat %>% 
    filter(ModuleColors == module.color) %>%
    dplyr::select(GeneSymbol, all_of(module.colname))
}
lapply(WGCNA_dat_select, nrow)
```

### Overrepresentation Analysis

```{r WGCNA ora}
devtools::source_url("https://github.com/TranLab/gene-set-functions/blob/main/ApplyORA2Genesets.R?raw=TRUE")

myuniverse <- unique(fData(x)[fData(x)$EnsemblID %in% colnames(WGCNA_matrix),]$GeneSymbol)
ORA_baseline_bound_df <- c()
minSize <- 15

#Combine similar modules skyblue1, mediumpurple1, thistle3 in to one
WGCNA_dat_select2 <- WGCNA_dat_select
names(WGCNA_dat_select2) <- mytopmodules$GeneSymbol
WGCNA_dat_select2$BTLA_CD22 <- data.table::rbindlist(list(WGCNA_dat_select2$BTLA,
                                                          WGCNA_dat_select2$CD22), use.names = FALSE) %>%
  as.data.frame() %>%
  rename(kMEMEall = "kMEMEpaleturquoise")
WGCNA_dat_select2$RIOK3_SEC62_TRIP12 <- data.table::rbindlist(list(WGCNA_dat_select2$RIOK3,
                                                                   WGCNA_dat_select2$SEC62,
                                                                   WGCNA_dat_select2$TRIP12), use.names = FALSE) %>%
  as.data.frame() %>%
  rename(kMEMEall = "kMEMEmidnightblue")


for(k in c("KDM5D","NKG2-E", "BTLA_CD22", "RIOK3_SEC62_TRIP12")){
  WGCNA_dat_select2[[k]] <- WGCNA_dat_select2[[k]][order(WGCNA_dat_select2[[k]][,2], decreasing = TRUE),]
  ORA_baseline_bound_df[[k]] <- ApplyORA2Genesets(genelist = WGCNA_dat_select2[[k]]$GeneSymbol,
                                                  geneset = "all",
                                                  universe = myuniverse,
                                                  output_directory = tempdir(),
                                                  filename_prefix = paste0("ORA_Mod_Corr_Protect_3_mos_", k,
                                                                           "_minSize", minSize),
                                                  minSize = minSize)
  }


KDM5D_ORA_baseline_bound <- bind_rows(ORA_baseline_bound_df$KDM5D, .id = "module_type") %>%
  mutate(module = "coral4") %>%
  mutate(module_size = nrow(WGCNA_dat_select2$KDM5D)) %>%
  mutate(hub_gene = "KDM5D")

NKG2E_ORA_baseline_bound <- bind_rows(ORA_baseline_bound_df$`NKG2-E`, .id = "module_type") %>%
  mutate(module = "lightblue4") %>%
  mutate(module_size = nrow(WGCNA_dat_select2$`NKG2-E`)) %>%
  mutate(hub_gene = "NKG2-E")

BTLA_CD22_ORA_baseline_bound <- bind_rows(ORA_baseline_bound_df$BTLA_CD22, .id = "module_type") %>%
  mutate(module = "the_two_amigos") %>%
  mutate(module_size = nrow(WGCNA_dat_select2$BTLA_CD22)) %>%
  mutate(hub_gene = "BTLA_CD22")

RIOK3_SEC62_TRIP12_ORA_baseline_bound <- bind_rows(ORA_baseline_bound_df$RIOK3_SEC62_TRIP12, .id = "module_type") %>%
  mutate(module = "the_three_amigos") %>%
  mutate(module_size = nrow(WGCNA_dat_select2$RIOK3_SEC62_TRIP12)) %>%
  mutate(hub_gene = "RIOK3_SEC62_TRIP12")


ORA_baseline_allbound <- bind_rows(KDM5D_ORA_baseline_bound,
                                  NKG2E_ORA_baseline_bound,
                                  BTLA_CD22_ORA_baseline_bound,
                                  RIOK3_SEC62_TRIP12_ORA_baseline_bound) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  mutate(pct_overlap = 100*(overlap/module_size)) %>%
  dplyr::select(module, module_size, hub_gene, module_type, pathway, overlap,
                pct_overlap, size, overlapGenes, pval, padj, neglogpadj)
```

### Plot WGCNA Over-representation Results

baseline analysis

```{r plotdat wgcna ORA, warning=FALSE, message=FALSE}
myModuleTypes <- c("MSigDB_Hallmark_v7.4", "MSigDB_C2_kegg_v7.4", "highBTMs", "lowBTMs", "BloodGen3Module", "MonacoModules")
myORAClusterPlotDat <- ORA_baseline_allbound %>%
  mutate(pathway = gsub("VS", "v", pathway)) %>%
  mutate(pathway = gsub("Vd", "Vδ", pathway)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = sub(".*?\\_", "", pathway)) %>%
  mutate(pathway = fct_reorder(pathway, neglogpadj))  %>%
  arrange(desc(neglogpadj))%>%
  mutate(TextLabelColor = ifelse(module_type == "BloodGen3Module", scales::muted("red"),
                                 ifelse(module_type == "MSigDB_C2_kegg_v7.4", scales::muted("blue"),
                                        ifelse(module_type == "MSigDB_Hallmark_v7.4", "black","gray")))) %>%
  filter(padj < 0.01) %>%
  filter(pct_overlap >= 5) %>%
  filter(module_type %in% myModuleTypes) %>%
  group_by(hub_gene) %>%
  arrange(neglogpadj) %>%
  slice_tail(n = 4) %>%
  ungroup()
```

### Baseline WGCNA Top Modules ORA Bubble plots (Figure S5B)
 
Over-representation analysis of genes within indicated networks which were determined by
weight gene correlation network analysis (WGCNA) of whole-blood Δ transcriptomes.

```{r prepare , warning=FALSE, message=FALSE, fig.width=6, fig.height=6}
addSmallLegend <- function(myPlot, pointSize = 2.5, textSize = 4, spaceLegend = 0.5) {
    myPlot +
        guides(shape = guide_legend(override.aes = list(size = pointSize)),
               color = guide_legend(override.aes = list(size = pointSize))) +
        theme(legend.title = element_text(size = textSize), 
              legend.text  = element_text(size = textSize),
              legend.key.size = unit(spaceLegend, "lines"))
}

myfont <- "Helvetica"
bubble_max_size <- 10
basetextsize <- 8
colorlabels <- myORAClusterPlotDat$TextLabelColor
myORAClusterPlot <- myORAClusterPlotDat %>%
  ggplot(., aes(x = hub_gene, y = pathway)) +
  geom_point(aes(size=neglogpadj, fill = pct_overlap), alpha = 0.65, shape=21, stroke = 0.25) +
  scale_size_area(name = expression(-log[10]~adj.~p~value), max_size = bubble_max_size) +
  viridis::scale_fill_viridis(option= "A", begin = 0.25, end = 0.75, alpha = 0.8, direction = -1, name = "% module overlap") +
  hrbrthemes::theme_ipsum_es(base_family = myfont, base_size = basetextsize) +
  scale_x_discrete(limits=rev) +
  scale_y_discrete(position = "right") +
  ylab("pathway/module") +
  xlab("network hub gene(s)") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 1, hjust=0)) +
  coord_flip()
```


```{r plot delta bubble, fig.align='center', fig.height=5, fig.width=6, echo=FALSE}
# Apply on original plot
addSmallLegend(myORAClusterPlot)
```

```{r save plot delta bubble, fig.align='center', fig.height=4.5, fig.width=5, echo=FALSE, eval=FALSE}
# Apply on original plot
# cairo_pdf(filename = "/Users/tuantran/OneDrive - Indiana University/FromGoogle/Manuscripts/KSPZV1 Manuscript/JCI Submission November 2022/Figures for JCI submission/Fig S5D delta vax WGCNA ORA.pdf", width = 5, height =4)
# addSmallLegend(myORAClusterPlot)
# dev.off()
```

```{r FGSEA on WGCNA Results, warning=FALSE, message=FALSE, eval=FALSE, echo=FALSE}
#FGSEA on WGCNA Results
#rank by module membership
minSize <- 5
# Make rank list; ranked by ModuleMembership correlation
  ranks <- GSEA_baseline_bound_df <- c()
devtools::source_url("https://github.com/TranLab/ModuleLists/blob/main/NamedGeneRankList2GseaTable.R?raw=TRUE")
for(k in names(WGCNA_dat_select)){
  WGCNA_dat_select[[k]] <- WGCNA_dat_select[[k]][order(WGCNA_dat_select[[k]][,2], decreasing = TRUE),]
  ranks[[k]] <- WGCNA_dat_select[[k]][,2]
  names(ranks[[k]]) <- WGCNA_dat_select[[k]]$GeneSymbol
  GSEA_baseline_bound_df[[k]] <- NamedGeneRankList2GseaTable(rankedgenes = ranks[[k]], geneset = "all", output_directory = tempdir(),
                                                        filename_prefix = paste0("FGSEA_Mod_Corr_Protect_3_mos_", k,
                                                                                 "_minSize", minSize), scoreType = "pos", minSize = minSize, fixed_seed = TRUE)
  }
```

```{r prepare data wgcna gsea, warning=FALSE, message=FALSE, eval=FALSE, echo=FALSE}
#Results
#delta analysis
myModuleTypes <- c("MSigDB_Hallmark_v7.4", "MSigDB_C2_kegg_v7.4", "highBTMs", "lowBTMs", "MonacoModules", "BloodGen3Module")
myGSEAClusterPlotDat <- GSEA_baseline_bound_df$MElightgreen %>%
  mutate(module_hub = "CD22") %>%
  bind_rows(., GSEA_baseline_bound_df$MEpaleturquoise %>%
              mutate(module_hub = "BTLA") ) %>%
  bind_rows(., GSEA_baseline_bound_df$MElightblue4 %>%
                mutate(module_hub = "NKG2-E")) %>%
  bind_rows(., GSEA_baseline_bound_df$MEcoral4 %>%
              mutate(module_hub = "KDM5D")) %>%
  bind_rows(., GSEA_baseline_bound_df$MEsalmon2 %>%
              mutate(module_hub = "TRIP12")) %>%
  bind_rows(., GSEA_baseline_bound_df$MEbrown4 %>%
              mutate(module_hub = "SEC62")) %>%
  bind_rows(., GSEA_baseline_bound_df$MEmidnightblue %>%
              mutate(module_hub = "RIOK3")) %>%
  mutate(neglogpadj = -log10(padj)) %>%
  mutate(module_hub = gsub("2-E", "2E", module_hub)) %>%
  mutate(pathway = gsub("VS", "v", pathway)) %>%
  mutate(pathway = gsub("Vd", "Vδ", pathway)) %>%
  mutate(pathway = gsub("gd", "γδ", pathway)) %>%
  mutate(pathway = sub(".*?\\_", "", pathway)) %>%
  mutate(pathway = fct_reorder(pathway, NES, .desc = TRUE))  %>%
  dplyr::arrange(desc(NES)) %>%
  mutate(TextLabelColor = ifelse(module_type == "lowBTMs", scales::muted("red"),
                                 ifelse(module_type == "highBTMs", scales::muted("blue"),
                                        ifelse(module_type == "BloodGen3Module", "orange",
                                               ifelse(module_type == "MonacoModules", "black",
                                                      ifelse(module_type == "highBTMs", scales::muted("blue"), 
                                                             ifelse(module_type == "MSigDB_C2_kegg_v7.4", scales::muted("green"),
                                                                    ifelse(module_type == "MSigDB_C2_kegg_v7.4", scales::muted("purple"), "gray")))))))) %>%
  filter(padj < 0.20) %>%
  filter(module_type %in% myModuleTypes) %>%
  droplevels()
```
